Effects of blood meal source and seasonality on reproductive traits of Culex quinquefasciatus (Diptera: Culicidae)

Version 1

Abstract

Abstract

Host selection by mosquitoes is a keystone to understand viral circulation and predict future infection outbreaks. Culex mosquitoes frequently feed on birds during spring and early summer, shifting into mammals towards late summer and autumn. This host switch may be produced by changes in mosquito fitness. The aim of this study was to assess if blood meal source and seasonality may influence reproductive traits of Culex quinquefasciatus mosquitoes. For this purpose, Cx. quinquefasciatus mosquitoes were reared in simulated summer and autumn conditions and fed with two different hosts, chickens and mice, in a factorial design. Fecundity, fertility and hatchability during two consecutive gonotrophic cycles were estimated. We found a greater fecundity and fertility for mosquitoes fed upon birds than mammals. Fecundity and fertility increased in autumn for chicken-fed mosquitoes, whereas they decreased for mouse-fed mosquitoes. These traits decreased in the second gonotrophic cycle for mouse-fed mosquitoes, whereas they did not vary between cycles for chicken-fed mosquitoes. Blood meal source had a significant effect on hatchability, finding the lowest hatchability in mouse-fed mosquitoes during the second gonotrophic cycle, whereas no differences were detected among the remaining treatments. These results indicate that fecundity and fertility are influenced by blood meal source and seasonality, increasing towards autumn for mosquitoes fed on birds and decreasing for mosquitoes fed on mammals. Hatchability seems also being influenced by blood meal source. Our results suggest that blood meal source and seasonality induce metabolic changes in the mosquito reproductive physiology which might affect host selection patterns.

Introduction

The southern house mosquito, Culex quinquefasciatus Say (Diptera: Culicidae), is a worldwide distributed vector of several pathogens, such as Wuchereria bancrofti Cobbold (Kasili et al., 2009), Dirofilaria immitis Leidy (Labarthe et al., 1998) and avian malaria (Farajollahi et al., 2011). Besides, this mosquito is also one of the main responsible for the transmission of the arboviruses West Nile virus (WNV) and Saint Louis encephalitis virus (SLEV), among others (Farajollahi et al., 2011). Because of this, this species is of critical concern due to its impact on both public and veterinary health.

Since arboviruses are transmitted by the bite of infectious mosquitoes between vertebrate hosts, activity of these pathogens is determined, at least in part, by the blood feeding habits of these vectors. Host selection of Culex mosquitoes is variable and it can be classified in four categories: (1) mammophilic (Muturi et al., 2008), (2) ornithophilic (Jansen et al., 2009), (3) herpetophilic (Janssen et al., 2015) and (4) generalist (Molaei et al., 2007). Moreover, some species may experience a host switch in its feeding habits, from avian to mammal hosts (Hancock & Camp, 2022). These species are of greater concern from an epidemiological standpoint because they can act as bridge vectors in the transmission of some arboviruses, such as SLEV and WNV (Kilpatrick et al., 2005).

Culex quinquefasciatus is considered an ornithophilic species in several areas around the globe (Takken & Verhulst, 2013). Despite its ornithophilic nature, Cx. quinquefasciatus and other Culex species could feed primarily from mammal hosts, including the human, in certain situations. Different studies have pointed out that in autumn there is an increased activity of SLEV and WNV in human populations (Kilpatrick et al., 2006, Spinsanti et al., 2008), suggesting that viral activity spills over to mammals from birds as a result of the mosquito feeding host switch (Edman & Taylor, 1968, Kilpatrick et al., 2006). A few hypotheses have attempted to explain why primarily ornithophilic mosquitoes also feed upon mammals in some contexts. One of the most widely accepted hypothesis states that the shift on feeding behavior of Culex mosquitoes is a consequence of the autumn migration of American robin (Turdus migratorius), the principal host in the northeastern US, generating an opportunistic change into mammal hosts (Kilpatrick et al., 2006). However, because this avian species is lacking in Southern latitudes it is an implausible hypothesis for many austral areas, being restricted only for some regions of the US. A second hypothesis proposed by Burkett-Cadena et al. (2011) ensures that host breeding cycles drive the host shift of mosquitoes, and peaks of host use are detected in periods of reproductive investment. According to this hypothesis, in hosts with parental cares, during reproductive seasons there is a greater investment of energy in assuring the offspring survivorship, consequently producing an increase in susceptibility of being bitten by mosquitoes as a result of a decrease in defensive behaviors (Burkett-Cadena et al., 2011). Other hypotheses suggest that mosquito host shift is a result of changes in host preference throughout the seasons (Edman, 1974), changes in density and feeding success of mosquitoes (Nelson et al., 1976) or seasonal changes in host habitat use by mosquitoes, however, they were poorly or not supported by field data (Edman, 1974; Burkett-Cadena et al., 2011).

Several biological (stress, metabolic rate, blood meal source) and environmental (temperature, photoperiod, i.e., season) variables that influence mosquito physiology might have an effect on reproduction traits, such as fecundity, development rate and survivorship, generating new nutritional requirements (Ciota et al., 2014, Costanzo et al., 2015, Gervasi et al., 2016; Yan et al. 2017, 2018). These factors, ultimately, could produce seasonal variation in host use by mosquitoes. The aim of the present study is to evaluate the effect of both host blood meal source and seasonality (combination of temperature and photoperiod) on three reproductive traits (fecundity, fertility and hatchability) of Culex quinquefasciatus mosquitoes. We hypothesize that the seasonal host switch of mosquitoes is the result of the effect of blood meal source and seasonality on reproductive outputs. We expect a greater number of eggs (fecundity) and larvae (fertility) in mosquitoes fed on mammal in autumn than in an avian host, expecting the opposite pattern in summer.

Materials and methods

Establishment and rearing of mosquito colony

Egg rafts of Culex quinquefasciatus were collected from a drainage ditch at Universidad Nacional de Córdoba Campus, Córdoba city in February 2021. Each raft was maintained individually in plastic containers with one liter of distilled water. The hatched larvae were fed with 100 mg of liver power three times per week until pupation. Pupae were transferred to a plastic emerging cage (21 cm x 12 cm) covered with a tulle-like fabric, containing distilled water and without food. Adults emerging from each raft were identified morphologically (Darsie, 1985) and molecularly (Smith & Fonseca, 2004) to make sure they corresponded to Cx. quinquefasciatus since in Córdoba is also present Cx. pipiens and its hybrids (Branda et al., 2021). All adults were reared in a cardboard cage of 22.5 liters (28 cm x 36.5 cm) and fed ad libitum with a 10% of sugar solution soaked in cotton pads on plastic glasses. For long-term maintenance of the colony, 24-h starved mosquitoes were offered with a blood meal of a restrained chicken twice a month. Four days post-feeding, a plastic recipient with distilled water was located inside the cage to allow engorged females lay egg rafts. Batches were collected and transferred to plastic containers (30 cm x 25 cm x 7 cm) in a proportion of 3 rafts per container, filled with 3 liters of distilled water. The hatched larvae were also fed with liver power at the same proportion described above. Pupae were transferred to emerging cages and adults to the final cardboard cage. The colony was inbred and maintained during several generations (generation > F20) in the Insectary of the Instituto de Virología “Dr. J.M. Vanella” (InViV). Room controlled conditions were 28°C, 12L:12D photoperiod and 70% relative humidity.

Experimental design

Blood source

For experimental trials, an avian and mammal model was used to evaluate the effect of blood meal source on fecundity, fertility and hatchability. The former host was represented by live chicks from the species Gallus gallus and the latter by live mice of the species Mus musculus (strain C57BL/6). For each trial, 24-h starved adult female mosquitoes were provided with a blood meal from restrained live chicks or mice. Chicks were donated by the Bartolucci poultry farm (Córdoba, Argentina). Mice were commercially obtained at the Instituto de Investigación Médica Mercedes y Martín Ferreyra (CONICET - Universidad Nacional de Córdoba). Vertebrate hosts were offered 1 hour before the lights went out and kept for 3 hours during 2 consecutive gonotrophic cycles.

The experimental use of animals (Gallus gallus and Mus musculus) was approved by the ethical committee at Facultad de Ciencias Médicas, Universidad Nacional de Córdoba (FCM-UNC) in compliance with the legislation regarding the use of animals for experimental and other scientific purposes (accession code: CE-2022-00518476-UNC-SCT#FCM).

Seasonality

To assess the effect of seasonality (photoperiod + temperature) on fecundity, fertility and hatchability, a “typical” summer and autumn day from Córdoba city were simulated in an incubator where mosquitoes were kept throughout the entire experiment. Summer conditions were: T°min = 22°C, T°max = 28°C, photoperiod = 14L:10D. Autumn conditions were: T°min = 16°C, T°max = 22°C, photoperiod = 10L: 14D. H umidity: 60-70% for both conditions. Data were obtained from Climate Data website (https://es.climate-data.org/)

Feeding trial design

Four experimental colonies were established from egg rafts collected from the maintenance colony, based on the combination of blood source and seasonality: bird-summer, bird-autumn, mammal-summer and mammal-autumn. These experimental colonies were reared at controlled room conditions from eggs to adults to ensure all adults have the same size (Kauffman et al., 2017), then they were transferred to the incubator to begin the assay. Adults were kept in cardboard cages of 8.3 liters (21 cm x 24 cm), fed ad libitum with a 10% of sugar solution soaked in cotton pads on plastic glasses.

Experiments were performed during two consecutive blood meals at 5 and 14 days post-emerging. After each blood meal, female mosquitoes were anesthetized with CO2 and classified as full engorged (VI Sella’s stage), partially fed (I-V Sella’s stage) and unfed (I Sella’s stage) (Silva Santos et al., 2019). Full engorged females were counted and separated into another cage to complete its gonotrophic cycle. Four days post-feeding a glass with distilled water was located inside the cage to allow oviposition. After each oviposition cycle egg rafts were collected, counted, and transferred to a 12-well plate with 3 mL of distilled water, where they were photographed to posteriorly count the number of eggs per raft. Rafts were maintained in each well until they hatched into L1 larvae, time where they were fixed with 1 mL of 96% ethanol. After hatching, the number of L1 larvae per raft was counted.

Statistical analysis

The reproductive outputs measured for each experimental colony and gonotrophic cycle were fecundity, fertility and hatchability. All analyses were performed using R Studio statistic software, v.4.2.1 (R Core Team, 2022). Fecundity was defined as the number of eggs per raft and fertility as the number of L1 larvae hatched per raft. To evaluate the effect of blood meal source and seasonality on these two variables, a Generalized Linear Model (GLM; package MASS) with negative binomial error distribution and logarithmic link function was adjusted (Venables & Ripley, 2002), with fecundity and fertility as response variables, respectively. Blood source, seasonality and gonotrophic cycle in a three-way interaction were set as explanatory variables. Hatchability (= hatching rate) was defined as the ratio of the number of larvae to the number of eggs per raft. Because data did not follow any known exponential family probability distribution (needed for linear regression) a null model analysis via data randomization was performed. Prior to this analysis we adjusted a factorial ANOVA using hatchability as response variable and blood source, seasonality and gonotrophic cycle in a three-way interaction as explanatory variables. From this model we obtained and saved the observed F value (FO) for each variable (main effects) and its combination (interaction effects). Posteriorly, we randomly permuted the response variable 10000 times and obtained simulated F value to create an own distribution for the F parameter. The FO value was compared to the simulated distribution to test the null hypothesis that data were generated randomly against the alternative that there might be a pattern generated for at least one of the explanatory variables.

Results

Effect of blood source and seasonality on fecundity and fertility

Blood meal source had a significant effect on fecundity but not in fertility. Seasonality had a significant effect on both fecundity and fertility. Gonotrophic cycle did not show an effect on neither of the two response variables (Tables 2, 3).

It was observed an interaction effect between blood meal source and seasonality on fecundity and fertility (Table 2, 3). For both response variables, it was not found differences in the number of eggs and larvae per raft in summer for both blood meal sources, chicken and mouse. However, in autumn it was detected an increase in fecundity and fertility for bird-fed mosquitoes and a decrease for mouse-fed mosquitoes (Fig. 1 A, C, Table 1). In general, avian-fed mosquitoes presented greater fecundity and fertility than mammal-fed mosquitoes, regardless of the season. However, from summer to autumn, it was observed an increase of 13% and 18% on fecundity and fertility, respectively, for chicken-fed mosquitoes and a decrease of 17% and 19%, respectively, for mouse-fed mosquitoes.

Interaction plots showing the marginal effects of blood meal source-seasonality and blood meal source-gonotrophic cycle on (A, B) fecundity and (C, D) fertility. Predicted mean values and its corresponding confidence intervals are shown. Means sharing same letters are not statistically different.

It was also detected an interaction effect between blood meal source and gonotrophic cycle (Table 2, 3). It was observed a decrease of 29% in fecundity and fertility from the first to second cycle for mouse-fed mosquitoes, being this treatment different from the remaining groups, in terms of fitness (Fig. 1 B, D, Table 1). During the second gonotrophic cycle fecundity and fertility for chicken-fed mosquitoes was 41% and 40% greater, respectively, than mosquitoes fed upon mice. It was not found statistically significant differences among the other three groups.

The rest of single factors and interactions from the fecundity and fertility models were not statistically significant (Table 2, 3).

Effect of blood source and seasonality on hatchability

By visual inspection it was detected that mosquitoes fed upon mammals possessed the greater hatchability, except for mosquitoes feeding on mammals in autumn during the second gonotrophic cycle, which registered the least hatching rate (Fig. 2). Statistically, it was detected a significant effect of blood meal source on hatching rate (FO = 0.00074, p = 0.0004). Highest hatchability values were observed in mouse-fed mosquitoes, ranging from 82% (autumn, second gonotrophic cycle) to 96% (autumn, first gonotrophic cycle) (Table 1, Fig. 2).

Boxplots showing the effects of both blood meal source and seasonality on hatchability for the first and second gonotrophic cycle.

Discussion

Several host physiological factors have been studied as putative candidates to explain the changes in mosquito feeding behavior, including stress, health status or metabolic rate (Gervasi et al., 2016; Yan et al. 2017, 2018). However, it has not been investigated how seasonal use of hosts affects reproductive traits in mosquitoes. It is known that some factors like host blood source, temperature or photoperiod have an effect on some reproductive aspects (Takken & Verhulst, 2013). Some authors have studied and proved the significant effects of blood meal source (Telang & Skinner, 2019), temperature (Ciota et al., 2014) and photoperiod (Costanzo et al., 2015) on fecundity and fertility. These studies illustrate how single effects of blood source and environmental factors may influence reproductive outputs. However, to our knowledge, there are no available studies that have attempted to assess the interaction effect among these factors, and most articles tend to investigate the effects of blood meal source and temperature or photoperiod independently.

Overall, we detected that chicken-fed mosquitoes produced more eggs and larvae per raft than mouse-fed mosquitoes, except in summer where differences were not detected. Akoh et al. (1993) reported a greater fecundity for mosquitoes fed on chicken than those fed on mice. Richards et al. (2012) and Telang & Skinner (2019) found a greater fecundity and fertility in chicken-fed mosquitoes than fed on bovine and pig blood, respectively. This effect is also present in other species of Culex (Shroyer & Silvery, 1972) and Aedes (Harrison et al., 2021). However, other authors have found the opposite pattern with human-fed mosquitoes (Ferdousi & Islam, 2005) and Guinea pig and human-fed mosquitoes (Xue et al., 2009) compared with mosquitoes fed upon chickens. In summary, with some few exceptions, mosquitoes fed on avian blood commonly produce a considerably greater batch size and fertility than mammal-fed mosquitoes. We also observed an increase of egg production and hatching per raft from summer to autumn in mosquitoes fed on chicks, whereas we found a decrease of these reproductive traits in mosquitoes fed upon mice. Regarding gonotrophic cycle, the second cycle was less productive in terms of fecundity and fertility for mosquitoes feeding on mice, however, we did not find differences in these reproductive outputs for mosquitoes fed on chicken. For hatchability, it was observed a greater hatching rate in mouse-fed mosquitoes than chicken-fed mosquitoes, however, the effect was not such marked as in fecundity and fertility.

Several hypotheses (-I) nutritional hypothesis, -II) digestive physiology hypothesis and -III) digestive mechanical hypothesis) have been proposed to explain why avian blood produces a greater fecundity and fertility even in mosquitoes with a known mammal-preferred host choice. The first widely accepted general idea (I) suggests that since birds have larger and nucleated erythrocytes, they provide greater amounts of proteins and energy than the anucleated mammalian erythrocytes, inducing a higher fecundity and fertility in mosquitoes feeding from these animals (Alto et al., 2014). However, it has been shown that mammalian blood contains more total protein, hemoglobin and hematocrit levels than avian blood, suggesting it is nutritional richer. Even though, mosquitoes such as Cx. quinquefasciatus experiments a process known as pre-diuresis, excreting large amounts of water (derived from blood serum), allowing it to concentrate and accumulate more proteins. Hence, it could be possible that mosquitoes compensate the low protein quality of bird blood by consuming larger volumes or feeding for longer times (Erram et al., 2022) which might explain the greater fecundity and fertility while feeding on avian hosts. More than the total protein, it has been shown that it is the amino acid balance which determines the egg production, and amino acid content varies between different animals. For Aedes aegypti, eight amino acids are required to ensure egg production, namely, arginine, isoleucine, leucine, lysine, phenylalanine, threonine, tryptophan and valine. Consumption of these amino acids induces the expression of yolk protein precursor genes and the resulting synthesis of the protein vitellogenin, which is absorbed by oocytes and allows the development of embryos and posterior egg laying. It has been demonstrated that from all these amino acids, isoleucine is the limiting one, and Greenberg (1951) found that addition of isoleucine to washed sheep erythrocytes increases egg production and oviposition. However, this effect was not observed with the addition of a mixture of the other amino acids. Since mammalian blood is poor in isoleucine compared with avian blood, it was suggested and proven for several species that isoleucine-rich host blood induces a greater production of eggs by mosquitoes (Greenberg, 1951; Harrison et al., 2021). The digestive physiology hypothesis (II) states that differences in egg production between avian and mammalian hosts is due to variation in digestion rates. Downe et al. (1963) found that in Mansonia perturbans digestion of avian blood occurred more rapidly than mammalian blood and this pattern was also confirmed in Culex tarsalis (Downe and Archer, 1975). In Aedes aegypti, nutrients from a blood meal can enter the oocyte with a period of 36-48 hs, after that the pinocytotic activity of the gamete decreases and the intake of exterior material is blocked. Because of this, those species of mosquitoes that can achieve the digestion of blood in a short period will ensure the exploitation of nutrients generating a greater egg production. This could be the situation present in mosquitoes fed upon birds, which produce a greater fecundity (Downe & Archer, 1975). A third possible explanation is the digestive mechanical hypothesis (III). Mosquitoes need to retain erythrocytes in midgut and hemolyzed the blood meal to exploit proteins. Female mosquitoes are armed with a set of sclerotized teeth-like structures present in the anterior foregut (= cibarial armature) and posterior midgut (= pyloric armature), the first one acts as the primary site of hemolysis, whereas the second armature prevents blood to escape outside the midgut, enhancing the exploitation of resources from red blood cells. Since the structure of cibarial and pyloric armature varies between mosquitoes and because erythrocytes from host are of different shapes and sizes, these factors could contribute to mosquitoes to select those hosts whose blood is easier to hemolyze and its protein easier to assimilate (Lyimo & Ferguson, 2009; Vaughan et al., 1991). In summary, several factors regarding blood quality and digestive physiology make bird blood a higher nutritional resource for egg production, and despite these factors were usually studied independently, it must be the combination or interaction of all of them which enhance the production of a greater fitness in mosquitoes feeding on birds.

Regarding environmental conditions, there is plentiful information available about the effects of temperature and photoperiod on physiological parameters (survival rate, development time, fecundity, fertility, behavior, etc.) (Hoffman, 1985). In nature, mosquitoes are exposed to a cyclic variation of both temperature and day length. It can be recognise a warm phase (= thermophase) which coincides with a photophase (light presence) and a cool phase (= cryophase) coinciding with a scotophase (light absence) (Joshi, 1996). Field and laboratory data have established that all these physiological traits follow a unimodal or hump-shaped relationship in ectothermic organisms, increasing from zero at a thermal minimum to a highest optimum and then decreasing to zero again at a thermal maximum. The best performance of animals generally is obtained at a narrow “optimal range”, which varies among and within species and traits. In Culex quinquefasciatus the fecundity curve ranges from a thermal minimum of 5°C to a maximum of 37.7°C with an optimal temperature of 21.4°C (Mordecai et al., 2019). However, the greatest egg production is restricted to a narrow interval around the thermal optimum. This narrow band has been determined for some species, for Culex pipiens complex and Aedes aegypti it seems that the optimal temperature ranges from 20°C to 30°C (Spanoudis et al., 2019; Bader et al. 2012).

The effect of temperature on physiological traits has been studied in two ways, keeping it constant across all the experiments or fluctuating it in different ways, generally between a minimum and maximum. For different populations of Culex quinquefasciatus and for a constant temperature, Mogi (1992) has found that fecundity increases from 15°C to 20°C and decreases to 28°C. Same pattern was registered for Aedes albopictus, with a peak at 26°C and decreasing to 30°C. Fluctuating temperature has the same pattern, although because temperature is allowed to fluctuate within a range mosquitos seem to cope better the extreme conditions, for example, Aedes albopictus produce same quantity of eggs at 18°C-26°C and 22°C-30°C (Joshi, 1996). Regarding photoperiod, fewer studies are available, some of them are those of Mogi (1992) and Costanzo et al. (2015), however, the effects are not as evident as temperature, and while Mogi (1992) found that mid and long day lengths induced greater fecundity, Costanzo et al. (2015) did not find differences of day length on fecundity, then, data are more contradictory than the effects of temperature.

Two important mosquito traits that might affect fecundity and the subsequent fertility are calendar age (= age post-emerging) and physiological age (= gonotrophic cycle). In general, fecundity and fertility decline with calendar age (Akoh et al., 1993; McCann et al., 2009) and gonotrophic cycle (Richards et al., 2012; Telang & Skinner, 2019). This general pattern was shown by us in mouse-fed mosquitoes. Although we also recorded a numerical increase in fecundity and fertility between gonotrophic cycles for mosquitoes fed on chicken, this difference was not significant. Same pattern was found by Bennet (1970) in Aedes aegypti fed upon avian and mammalian hosts. In addition, Obholz (unpublished data) also found an increase in fecundity and fertility for Ae. Aegypti fed on mice. A difference in egg production may be attributable to variation in feeding rate, but since we only evaluated full-engorged female mosquitoes this difference cannot be explained based on this rate. Hence, there must be an underlying physiological or environmental process that may be affecting fecundity and fertility and further studies are needed to corroborate it.

All studies assessing the effect of temperature or photoperiod on reproductive traits have been independent from the source of blood meal, however, our results suggest that these two environmental variables have a significant effect on fecundity and fertility, affecting differentially regarding blood meal source. We found that in summer (mean temperature = 25°C and long-day photoperiod) there is not differences on fecundity and fertility for mosquitoes fed upon mouse and chicken, however, in autumn (mean temperature = 19°C and short-day photoperiod) the fecundity and fertility of chicken-fed mosquitoes increased whereas in mouse-fed mosquitoes decreased. These results are not directly comparable with literature, since there are no studies that have evaluated the interaction effect of blood meal source and seasonality (temperature + photoperiod). Several authors suggest that fecundity and fertility increase at higher temperatures and photoperiod for both bird and mammal-fed mosquitoes (Abdallah et al., 2020; Al-Rashidi et al., 2022, Shehata, 2018), however, a difference based on blood meal source is elusive. This little evidence makes it evident that more studies are needed to assess the combined effect of temperature, photoperiod and blood source on reproductive traits since a common pattern has not been readily identified at the moment.

Fertility in our experiments varied among treatments. Considering the number of eggs, differences in fertility may be attributable to two factors: (I) variation in fecundity or (II) variation in hatching rate (= hatchability). The first has already been tested, then, to assess if the variation in number of larvae was a result of differences in hatchability, we evaluated if there were changes in the percentage of eggs hatched. We find significant differences in hatchability regarding the type of blood meal source, obtaining a mean hatching rate for mammal-fed mosquitoes of 92% and 87% for bird-feeding mosquitoes. In literature, hatchability (also cited as viability or even fertility) has been shown to vary according to different temperatures (van der Linde et al., 1990) but not with blood meal sources (dos Santos Dias et al., 2018; Richard et al., 2012). Our preliminary data suggest that changes in fertility are a consequence of variation in egg production due to the effects of blood meal source and seasonality and not in hatchability

Understanding factors leading host selection and host switch are critical to predict arboviral outbreaks and arbovirus spill over. It is well known that many species of Culex mosquitoes experience changes in their host usage, from birds in spring and late summer to mammals in late summer and autumn, which usually trigger the increase of viral infection reported on humans, turning from a period of viral amplification to viral transmission. Nevertheless, studies considering the causes of host switching are very scarce and they mainly focus on host factors, such as host availability or defensive behavior (Kilpatrick et al., 2006, Burkett-Cadena et al., 2011). A variety of mosquito physiological traits have been studied as an attempt to explain host selection but neither of them have been applied in host switch mechanisms (Lyimo & Ferguson, 2009; Yan et al., 2017, 2018). In this work we investigated for the first time if some reproductive traits could explain the pattern in host shift from birds to mammals. In our study we included the interaction effects of blood meal source and seasonality on fecundity and fertility, two variables analyzed separately but never combined, to the best of our knowledge. Our results suggest that there is a change in fecundity and fertility through seasons due to the effects of blood meal source and seasonality.

In summary, our model suggests that biological and climatic factors induce physiological changes in mosquitoes which triggers selection of new hosts, generating an increase of fitness (fertility and fecundity). Our results did not support the hypothesis that greater fitness on birds is present in summer and on mammals in autumn, since we register the opposite pattern. Notwithstanding, since Culex mosquitoes have a wide range of host niche it remains inconclusive whether other factors may be intervening in mosquitoes seasonal host shift.

eLife Assessment

This useful study presents convincing evidence that blood meal source and season affect Culex quinquefasciatus mosquito reproduction. Its unique focus on the interactive effects of both factors on mosquito fitness is of considerable relevance for the field, but the work suffers from inadequate experimental design - no replication, a population mismatch with the hypothesis region, and small sample sizes, limitations that were not sufficiently acknowledged in the discussion. The work will be of interest to those studying malaria and vector-borne diseases.

Reviewer #1 (Public Review):

Summary:

This study examines the role of host blood meal source, temperature, and photoperiod on the reproductive traits of Cx. quinquefasciatus, an important vector of numerous pathogens of medical importance. The host use pattern of Cx. quinquefasciatus is interesting in that it feeds on birds during spring and shifts to feeding on mammals towards fall. Various hypotheses have been proposed to explain the seasonal shift in host use in this species but have provided limited evidence. This study examines whether the shifting of host classes from birds to mammals towards autumn offers any reproductive advantages to Cx. quinquefasciatus in terms of enhanced fecundity, fertility, and hatchability of the offspring. The authors found no evidence of this, suggesting that alternate mechanisms may drive the seasonal shift in host use in Cx. quinquefasciatus.

Strengths:

Host blood meal source, temperature, and photoperiod were all examined together.

Weaknesses:

The study was conducted in laboratory conditions with a local population of Cx. quinquefasciatus from Argentina. I'm not sure if there is any evidence for a seasonal shift in the host use pattern in Cx. quinquefasciatus populations from the southern latitudes.

Reviewer #2 (Public Review):

Summary:

Conceptually, this study is interesting and is the first attempt to account for the potentially interactive effects of seasonality and blood source on mosquito fitness, which the authors frame as a possible explanation for previously observed host-switching of Culex quinquefasciatus from birds to mammals in the fall. The authors hypothesize that if changes in fitness by blood source change between seasons, higher fitness in birds in the summer and on mammals in the autumn could drive observed host switching. To test this, the authors fed individuals from a colony of Cx. quinquefasciatus on chickens (bird model) and mice (mammal model) and subjected each of these two groups to two different environmental conditions reflecting the high and low temperatures and photoperiod experienced in summer and autumn in Córdoba, Argentina (aka seasonality). They measured fecundity, fertility, and hatchability over two gonotrophic cycles. The authors then used a generalized linear mixed model to evaluate the impact of host species, seasonality, and gonotrophic cycle on fecundity and fertility and a null model analysis via data randomization for hatchability. The authors were trying to test their hypothesis by determining whether there was an interactive effect of season and host species on mosquito fitness. This is an interesting hypothesis; if it had been supported, it would provide support for a new mechanism driving host switching. While the authors did report an interactive impact of seasonality and host species, the directionality of the effect was the opposite of that hypothesized. While this finding is interesting and worth reporting, there are significant issues with the experimental design and the conclusions that are drawn from the results, which are described below. These issues should be addressed to make the findings trustworthy.

Strengths:

1. Using a combination of laboratory feedings and incubators to simulate seasonal environmental conditions is a good, controlled way to assess the potentially interactive impact of host species and seasonality on the fitness of Culex quinquefasciatus in the lab.

2. The driving hypothesis is an interesting and creative way to think about a potential driver of host switching observed in the field.

Weaknesses:

1. There is no replication built into this study. Egg lay is a highly variable trait, even within treatments, so it is important to see replication of the effects of treatment across multiple discrete replicates. It is standard practice to replicate mosquito fitness experiments for this reason. Furthermore, the sample size was particularly small for some groups (e.g. 15 egg rafts for the second gonotrophic cycle of mice in the autumn, which was the only group for which a decrease in fecundity and fertility was detected between 1st and 2nd gonotrophic cycles). Replicates also allow investigators to change around other variables that might impact the results for unknown reasons; for example, the incubators used for fall/summer conditions can be swapped, ensuring that the observed effects are not artifacts of other differences between treatments. While most groups had robust sample sizes, I do not trust the replicability of the results without experimental replication within the study.

2. Considering the hypothesis is driven by the host switching observed in the field, this phenomenon is discussed very little. I do not believe Cx. quinquefasciatus host switching has been observed in Argentina, only in the northern hemisphere, so it is possible that the species could have an entirely different ecology in Argentina. It would have been helpful to conduct a blood meal analysis prior to this experiment to determine whether using an Argentinian population was appropriate to assess this question. If the Argentinian populations don't experience host switching, then an Argentinian colony would not be the appropriate colony to use to assess this question. Given that this experiment has already been conducted with this population, this possibility should at least be acknowledged in the discussion. Or if a study showing host switching in Argentina has been conducted, it would be helpful to highlight this in the introduction and discussion.

3. The impacts of certain experimental design decisions are not acknowledged in the manuscript and warrant discussion. For example, the larvae were reared under the same conditions to ensure adults of similar sizes and development timing, but this also prevents mechanisms of action that could occur as a result of seasonality experienced by mothers, eggs, and larvae.

4. There are aspects of the data analysis that are not fully explained and should be further clarified. For example, there is no explanation of how the levels of categorical variables were compared.

5. The results show the opposite trend as was predicted by the authors based on observed feeding switches from birds to mammals in the autumn. However, they only state this once at the end of the discussion and never address why they might have observed the opposite trend as was hypothesized.

6. Generally speaking, the discussion has information that isn't directly related to the results and/or is too detailed in certain parts. Meanwhile, it doesn't dig into the meaning of the results or the ways in which the experimental design could have influenced results.

7. Beyond the issue of lack of replication limiting trust in the conclusions in general, there is one conclusion reached at the end of the discussion that would not be supported, even if additional replicates are conducted. The results do not show that physiological changes in mosquitoes trigger the selection of new hosts. Host selection is never measured, so this claim cannot be made. The results don't even suggest that fitness might trigger selection because the results show that physiological changes are in the opposite direction as what would be hypothesized to produce observed host switches. Similarly, the last sentence of the abstract is not supported by the results.

8. Throughout the manuscript, there are grammatical errors that make it difficult to understand certain sentences, especially for the results.

This study is driven by an interesting question and has the potential to be a valuable contribution to the literature.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

This study examines the role of host blood meal source, temperature, and photoperiod on the reproductive traits of Cx. quinquefasciatus, an important vector of numerous pathogens of medical importance. The host use pattern of Cx. quinquefasciatus is interesting in that it feeds on birds during spring and shifts to feeding on mammals towards fall. Various hypotheses have been proposed to explain the seasonal shift in host use in this species but have provided limited evidence. This study examines whether the shifting of host classes from birds to mammals towards autumn offers any reproductive advantages to Cx. quinquefasciatus in terms of enhanced fecundity, fertility, and hatchability of the offspring. The authors found no evidence of this, suggesting that alternate mechanisms may drive the seasonal shift in host use in Cx. quinquefasciatus.

Strengths:

Host blood meal source, temperature, and photoperiod were all examined together.

Weaknesses: The study was conducted in laboratory conditions with a local population of Cx. quinquefasciatus from Argentina. I'm not sure if there is any evidence for a seasonal shift in the host use pattern in Cx. quinquefasciatus populations from the southern latitudes.

We agree on the reviewers observation about the evidence on seasonal shift in the host use pattern in Cx. quinquefasciatus populations from southern latitudes. We include a paragraph in the Introduction section regarding this. Unfortunately, studies conducted in South America to understand host use by Culex mosquitoes are very limited, and there are virtually no studies on the seasonal feeding pattern. In Argentina, there is some evidence (Stein et al., 2013, Beranek, 2019) regarding the seasonal change in host use by Culex species, including Cx. quinquefasciatus, where the inclusion of mammals during the autumn has been observed. As part of a comprehensive study on characterising bridge vectors for SLE and WN viruses, our research group is currently working on the molecular identification of blood meals from engorged females to gain deeper insights into the seasonal feeding pattern of Culex mosquitoes. While the seasonal change in host use by Culex quinquefasciatus has not been reported in Argentina so far, there has been an observed increase in reported cases of SLE virus in humans between summer and fall (Spinsanti et al., 2008). It is based on this evidence that we hypothesise there is a seasonal change in host use by Cx. quinquefasciatus, similar to what occurs in the United States. This is also considering that both countries (Argentina and the United States) have regions with similar climatic conditions (temperate climates with thermal and hydrological seasonality). Since we work on the same species and in a similar temperate climate regimen, we assumed there is a seasonal shift in the host use by this mosquito species.

Reviewer #1 (Recommendations for the authors):

Abstract

Line 23: fed on two different hosts.

Accepted as suggested.

I think the concluding statement should be rewritten to say that immediate reproductive outcomes do not explain the shift in host use pattern of Cx. quinquefasciatus mosquitoes from birds to mammals towards autumn.

Accepted as suggested.

Introduction

No comments.

Materials and Methods

Please mention sample sizes in the text as well (n = ?) for each treatment.

Accepted as suggested.

Page 99: ......C. quinquefasciatus, since C. pipiens and its hybrids are present as well in Cordoba.

Accepted as suggested.

Results – Line 146: subsequently instead of posteriorly

Accepted all changes as suggested.

Line 148: were counted instead of was counted.

Accepted all changes as suggested.

Line 160: Subsequently instead of posteriorly

Accepted all changes as suggested.

Line 171: on fertility

Accepted all changes as suggested.

Line 174: there was an interaction effect on…

Accepted all changes as suggested.

Line 175: there were no differences in the number of eggs

Accepted all changes as suggested.

Discussion

I think the first paragraph in the discussion section is redundant and should be deleted.

The whole discussion was rewritten to be focused on our aims and results.

Line 282: this sentence needs to be rewritten.

Accepted as suggested.

Line 299: at 28{degree sign}C

Line 300: at 30{degree sign}C

Sorry, but we are not sure about your comment here. We checked. Temperatures are written as stated, 28°C and 30°C.

Line 363: I think the authors need to discuss more about the bigger question they were addressing. I think that the discussion section can be strengthened greatly by elaborating on whether there is evidence for a seasonal shift in host use pattern in Cx. quinquefasciatus in the southern latitudes. If yes, what alternate mechanisms they believe could be driving the seasonal change in host use in this species in the southern latitudes now that they show the 'deriving reproductive advantages' hypothesis to be not true for those populations.

Thanks for this observation. We agree and so the Discussion section was restructured to align it with our results, as suggested.

Reviewer #2 (Public Review):

Summary:

Conceptually, this study is interesting and is the first attempt to account for the potentially interactive effects of seasonality and blood source on mosquito fitness, which the authors frame as a possible explanation for previously observed host-switching of Culex quinquefasciatus from birds to mammals in the fall. The authors hypothesize that if changes in fitness by blood source change between seasons, higher fitness in birds in the summer and on mammals in the autumn could drive observed host switching. To test this, the authors fed individuals from a colony of Cx. quinquefasciatus on chickens (bird model) and mice (mammal model) and subjected each of these two groups to two different environmental conditions reflecting the high and low temperatures and photoperiod experienced in summer and autumn in Córdoba, Argentina (aka seasonality). They measured fecundity, fertility, and hatchability over two gonotrophic cycles. The authors then used a generalized linear mixed model to evaluate the impact of host species, seasonality, and gonotrophic cycle on fecundity and fertility and a null model analysis via data randomization for hatchability. The authors were trying to test their hypothesis by determining whether there was an interactive effect of season and host species on mosquito fitness. This is an interesting hypothesis; if it had been supported, it would provide support for a new mechanism driving host switching. While the authors did report an interactive impact of seasonality and host species, the directionality of the effect was the opposite of that hypothesized. While this finding is interesting and worth reporting, there are significant issues with the experimental design and the conclusions that are drawn from the results, which are described below. These issues should be addressed to make the findings trustworthy.

Strengths:

(1) Using a combination of laboratory feedings and incubators to simulate seasonal environmental conditions is a good, controlled way to assess the potentially interactive impact of host species and seasonality on the fitness of Culex quinquefasciatus in the lab.

(2) The driving hypothesis is an interesting and creative way to think about a potential driver of host switching observed in the field.

Weaknesses:

(1) There is no replication built into this study. Egg lay is a highly variable trait, even within treatments, so it is important to see replication of the effects of treatment across multiple discrete replicates. It is standard practice to replicate mosquito fitness experiments for this reason. Furthermore, the sample size was particularly small for some groups (e.g. 15 egg rafts for the second gonotrophic cycle of mice in the autumn, which was the only group for which a decrease in fecundity and fertility was detected between 1st and 2nd gonotrophic cycles). Replicates also allow investigators to change around other variables that might impact the results for unknown reasons; for example, the incubators used for fall/summer conditions can be swapped, ensuring that the observed effects are not artefacts of other differences between treatments. While most groups had robust sample sizes, I do not trust the replicability of the results without experimental replication within the study.

We agree egg lay is a variable trait and so we consider high numbers of mosquitoes and egg lay during experiments compared to our studies of the same topics. Evaluating variables such as fecundity, fertility, or other types of variables (collectively referred to as "life tables") is a challenging issue that depends on several intrinsic and extrinsic factors. Because all of this, in some experiments, sample sizes might not be very large, and in several articles, lower sample sizes could be found. For instance, in Richards et al. (2012), for Culex quinquefasciatus, during the second gonotrophic cycle, some experiments had 13 or even 6 egg rafts. For species like Aedes aegypti, the sample size for life table analysis is also usually small. As an example, Muttis et al. (2018) reported between 1 and 4 engorged females (without replicates). In addition, small sample size would be a problem if we would not have obtained any effect, which is not the case due to the fact that we were interested in finding an effect, regardless of the effect size. Because of this, we do find our sample sizes quite robust for our results.

Regarding the need to repeat the experiments in order to give more robustness to the study we also agree. However, after a review of the literature (articles cited in the original manuscript), it is apparent that similar experiments are not frequently repeated as such. Examples of this are the studies of Richards et al. (2012), Demirci et al. (2014) or Telang & Skinner (2019), which even they manipulate several cages at a time as “replicates”, they are not true replicates because they summarise and manipulate all data together, and do not repeat the experiment several times. We see these “replicates” as a way of getting a greater N.

As was stated by the reviewer, repetition is a resource and time-consuming activity that we are not able to do. Replicating the experiment poses a significant time and resources challenge. The original experiment took over three months to complete, and it is anticipated that a similar timeframe would be necessary for each replication (6 months in total considering two more replicates). Given our existing commitments and obligations, dedicating such an extensive period solely to this would impede progress on other crucial projects and responsibilities.

Given the limitations of resources and time and the infrequent use of experimental replication in this type of studies, we performed a simulation-based analysis via a Monte Carlo approach. This approach involved generating synthetic data that mimics the expected characteristics of the original experiment and subsequently subjecting it to the same analysis routine. The main goal of this simulation was to evaluate the potential spuriousness and randomness of the results that might arise due to the experimental conditions. So, evaluating the robustness and confidence of our results and data.

(2) Considering the hypothesis is driven by the host switching observed in the field, this phenomenon is discussed very little. I do not believe Cx. quinquefasciatus host switching has been observed in Argentina, only in the northern hemisphere, so it is possible that the species could have an entirely different ecology in Argentina. It would have been helpful to conduct a blood meal analysis prior to this experiment to determine whether using an Argentinian population was appropriate to assess this question. If the Argentinian populations don't experience host switching, then an Argentinian colony would not be the appropriate colony to use to assess this question. Given that this experiment has already been conducted with this population, this possibility should at least be acknowledged in the discussion. Or if a study showing host switching in Argentina has been conducted, it would be helpful to highlight this in the introduction and discussion.

Thanks for this observation. We agree. However, we conducted the experiment beside host use data from Argentina since we used the mosquito species, and the centre region of Argentina (Córdoba) has a similar temperate weather regimen that those observed in the east coast of US.

We are aware that few studies regarding host shifting in South America are available, some such that those conducted by Stein et al. (2013) and Beranek (2019) reported a moderate host switch for Culex quinquefasciatus in Argentina. We have already performed a study about seasonal host feeding patterns for this species. However, even though there are few studies regarding host shifting, our hypothesis is based mainly in the seasonality of human cases of WNV and SLEV, a pattern that has been demonstrated for our region, see for example the study of Spinsanti et al. (2008).

We include a new paragraph in the Introduction and Discussion sections. Please see answers Reviewer #1.

(3) The impacts of certain experimental design decisions are not acknowledged in the manuscript and warrant discussion. For example, the larvae were reared under the same conditions to ensure adults of similar sizes and development timing, but this also prevents mechanisms of action that could occur as a result of seasonality experienced by mothers, eggs, and larvae.

We understand the confusion that may have arisen due to a lack of further details in the methodology. If we are not mistaken, you are referring to our oversight regarding the consideration of carry-over effects of larvae rearing that could potentially impact reproductive traits. When investigating the effects of temperature or other environmental factors on reproductive traits, it is possible to acclimate either larvae or adults. This is due to the significant phenotypic plasticity that mosquitoes exhibit throughout their entire ontogenetic cycle. In our study, we followed an approach similar to that of other authors where the adults are exposed to experimental conditions (temperature and photoperiod). For a similar approach you can refer to the studies conducted by Ferguson et al. (2018) for Cx. pipiens, Garcia Garcia & Londoño Benavides (2007) for Cx. quinquefasciatus or Christiansen-Jucht et al. (2014, 2015) for Anopheles gambiae.

(4) There are aspects of the data analysis that are not fully explained and should be further clarified. For example, there is no explanation of how the levels of categorical variables were compared.

The methodology and statistical analysis were expanded for a better understanding.

(5) The results show the opposite trend as was predicted by the authors based on observed feeding switches from birds to mammals in the autumn. However, they only state this once at the end of the discussion and never address why they might have observed the opposite trend as was hypothesized.

The discussion was restructured to focus on our results and our model.

(6) Generally speaking, the discussion has information that isn't directly related to the results and/or is too detailed in certain parts. Meanwhile, it doesn't dig into the meaning of the results or the ways in which the experimental design could have influenced results.

As mentioned above, the discussion was restructured to reflect our findings. We also included the effect that our design might have influenced our results. However, as stated above we do not fully agree that the design is inadequate for our analysis, we performed standard protocols followed by other researchers and studies in this research field.

(7) Beyond the issue of lack of replication limiting trust in the conclusions in general, there is one conclusion reached at the end of the discussion that would not be supported, even if additional replicates are conducted. The results do not show that physiological changes in mosquitoes trigger the selection of new hosts. Host selection is never measured, so this claim cannot be made. The results don't even suggest that fitness might trigger selection because the results show that physiological changes are in the opposite direction as what would be hypothesized to produce observed host switches. Similarly, the last sentence of the abstract is not supported by the results.

We agree with this observation. However, we did not evaluate the impact of fitness on host selection in this study. Instead, we aimed to investigate the potential influence of seasonality on mosquito fitness as a potential trigger for a shift in host selection. We agree that we have incorrectly used the term “host selection” when we should actually be discussing “host use change”. Our results indicate a seasonal alteration in mosquito fitness in response to temperature and photoperiod changes. Building upon this observation, we re-discussed our hypothesis and theoretical model to explain this seasonal shift in host use.

(8) Throughout the manuscript, there are grammatical errors that make it difficult to understand certain sentences, especially for the results.

All English grammar and writing of the manuscript was revised and corrected to be easily understood.

This study is driven by an interesting question and has the potential to be a valuable contribution to the literature.

Reviewer #2 (Recommendations for The Authors):

I hope that the authors will consider the suggested revisions and experimental replication to improve the quality of the study and paper.

This study tests a very interesting hypothesis. I understand that additional replicates are difficult to conduct, but I do believe that fitness studies absolutely require experimental replicates. Unless you are able to replicate the observed effects, I personally would not trust the results of this study. I hope that you will consider conducting replicates so that this important question can be answered in a more robust manner. Below, I expand upon some additional points in the public review and also provide more specific suggestions. I provided some copy-editing feedback, but was not able to point out all grammatical mistakes. I suggest that you use ChatGPT to help you edit the English. For example, you can feed ChatGPT your MS and ask it to bold the grammatical errors or you can ask it to edit grammatical errors and bold the sections that were edited. I understand that writing in a second language is very difficult (from personal experience!), so I view ChatGPT as a great tool to help even the playing field for publishing. Below are line item suggestions. Apologies that wording is curt, I was trying to be efficient in writing.

20-21: I suggest that you emphasize that you are investigating the interactive effect.

Accepted as suggested.

22: they weren't "reared" (from larvae) in different conditions, they were "maintained" as adults

Accepted as suggested.

26-27: increased/decreased is a bit misleading since you did not evaluate these groups sequentially in time. It might be more accurate to describe it as less than/greater than. Also, if you say increased/decreased or less than/greater than, you should always say what you are comparing to. The same applies throughout the MS.

Accepted as suggested.

29-30: "finding the" is not correct here; could be "with the lowest..."

Accepted as suggested.

34-36: I do not think that your results suggest this, even if you were to replicate the results of this experiment. You haven't shown metabolic changes.

We understand the point. Accepted as suggested.

42-44: "one of the main responsible" should be "one of the main species responsible..."

Accepted as suggested.

48: I think that "host preference" is better than selection here; -philic denotes preference

Accepted as suggested.

50: "Moreover" isn't the correct transition word here

Accepted as suggested.

57: "could" isn't correct here; consider saying "... species sometimes feed primarily on mammal hosts, including humans, in certain situations."

Accepted as suggested.

58: Different isn't correct word here

Accepted as suggested.

60: delete "feeding"

Accepted as suggested.

66-68: I am not familiar with any blood meal analysis studies in the southern hemisphere that show host switching for Culex species between summer and autumn. If this hasn't been shown, then this critique of the host migration hypothesis doesn't make sense.

There are some studies pointing this out (Stein et al., 2013, Beranek 2019), and unpublished data from us). However, our hypothesis has supported by epidemiological data observed in human population which indicate a seasonal activity pattern. It was explained in depth in the Introduction section.

68: ensures is not the right word; I suggest "suggests"

Accepted as suggested.

68-70: this explanation isn't clear to me; please revise

It will be revised. Accepted as suggested.

70: change cares to care

Accepted as suggested.

76-77: can you explain how they were not supported by the data for the benefit of those who are not familiar with these papers please?

Accepted as suggested.

87-89: I suggest the following wording: "In the autumn, we expect a greater number of eggs (fecundity) and larvae (fertility) in mosquitoes after feeding on a mammal host compared to an avian host, and the opposite relationship in the summer."

Accepted as suggested.

99: edit for grammar

Accepted as suggested.

102: suggest: "...offered a blood meal from a restrained chicken twice a month"

Accepted as suggested.

107: powder

Accepted as suggested.

108: inbred? Is this the term you meant to use?

Changed as suggested.

109: "several" cannot be used to describe 20 generations; suggest using "over twenty generations"; also, it would be good to acknowledge in your discussion that lab adaptation could force evolution, especially since mosquitoes are kept at constant temperatures and fed with certain hosts (with easy access) in the lab. Also, it would be good to know when the experiments were conducted to know the lapse of time between the creation of the colony and the experiments.

Accepted as suggested.

110-111: Does humidity vary between summer and fall in Córdoba? If so, I suggest acknowledging in the discussion that if humidity differences are involved in a potential interaction between host species and seasonality, then this would not have been captured by your experimental design.

Several variables change during seasons. We were interested in capturing the effects of temperature and photoperiod, since humidity is a variable difficult to control.

113-116: I suggest combining into one sentence to make more concise.

Accepted as suggested.

135: You might be obscuring the true impact of seasonality by rearing the larvae under the same conditions. There may be signals that mothers/eggs/larvae receive that influence their behavior (e.g. I believe this is the case for diapause), so this limitation should also be acknowledged. I understand why you decided to do this to control for development time and size, but it is something that should be considered in the discussion.

As it was explained above, Cx. quinquefasciatus do not suffer diapause in our country. Maintaining mosquitoes from adults was an approach selected by us based on other studies.

138: edit: "with cotton pads soaked in... on plastic..."; what is plastic glass? Do you mean plastic dishes?

Accepted as suggested.

141: here and throughout paragraph, full should be "fully"

Accepted as suggested.

144: located should be "placed"

Accepted as suggested.

147: suggest editing to "at which point, they were fixed with 1 mL of 96% ethanol and the number of L1 larvae per raft was counted."

Accepted as suggested.

154-155: edit for grammar

Accepted as suggested.

157: Your GLM explanation doesn't say anything about how you made pairwise comparisons between your levels; did you use emmeans?

This revised version includes a more detailed methodology and statistical analysis. Accepted as suggested.

158-160: I don't understand why you took this approach - it seems strange to me to use this analysis, but I am not familiar with it, so it might be that I lack the knowledge to be able to adequately evaluate. Please provide more explanation so that readers can better understand this analysis. A citation for this kind of application of the analysis would be helpful.

It was changed to be in accordance with the remaining analyses.

173: replace neither with either

Accepted as suggested.

174: this applies throughout; edit to : "An interaction effect was observed..."

Accepted as suggested.

175: "it was not found" is grammatically incorrect; instead : "We did not find ..." or "no differences in... were detected", etc

Accepted as suggested.

183: "it was detected" is grammatically incorrect

Accepted as suggested.

185-186: "being this treatment... in terms of fitness": I do not understand what this means. Please rephrase

Accepted as suggested.

170-199: you should provide the effect sizes and p values in text and/or in the figure for the pairwise comparisons

Accepted as suggested.

193-196. These two sentences are confusing and I am not sure what you mean, especially in the first sentence.

It was rewritten. Accepted as suggested.

Figure 1: This figure is great and easy to read and interpret! Thank you for the comment! 218-219: it is important to state which mosquito species you are referring to here.

Accepted as suggested.

226-227: you definitely should acknowledge the small sample size here.

Considered.

227: "it was observed" should be "We observed" or "A greater hatching rate.... was observed."

Accepted as suggested.

228-229: is the result really comparable even though you took very different approaches to the analysis for these outcomes?

Changed to be comparable.

230-278: the discussion of these hypotheses is too long and detailed, especially since the comparison of mouse vs chicken wasn't your main question; you really wanted to understand this in the context of seasonality. I suggest cutting this down a lot and making room to dig into your results more, and also to discuss the potential impacts of your experimental design/limitations on the results.

Discussion was changed to focus on our results and model. Accepted as suggested.

281: Hoffman is an old citation; I suggest you cite a modern review.

Accepted as suggested. We deleted it due to the re-writing of the manuscript.

282: "It can be recognise".. I am not sure what you are trying to say here

Accepted as suggested.

1. After the first time you write a species name, you can abbreviate the genus in all future mentions unless it is at the beginning of a sentence.

Accepted as suggested.

303-305: Revise this sentence. E.g "Fewer studies are available regarding photoperiod and show mixed results; Mogi (1992) found that mid and long day lengths induced greater fecundity while Costanzo et al. (2015) did not find differences in fecundity by day length."

Accepted as suggested.

315-316: typically, unpublished data shouldn't be referenced; I'm not sure if eLife has a policy on this.

We will check this with eLife guidelines. However, since the lack of evidence on this pattern we consider important to include this unpublished data.

316: Aegypti should be lowercase

Accepted as suggested.

328-330: This sentence is redundant with the first sentence of the paragraph

Accepted as suggested.

321-336: You never reintroduced your hypothesis in your discussion. I suggest that you center your whole discussion more directly around the hypothesis that motivated the study. If you decide not to restructure your discussion, you should at least reintroduce your hypothesis here and discuss how your results do not support the hypothesis.

Accepted as suggested.

337-348: This paragraph is a bit confusing as you jump between fertility and hatchability

Accepted as suggested.

353: is viral transmission the right word to use here? I think you might mean bridge vector transmission to humans specifically?

Accepted as suggested.

357: you say "neither" but never define which traits you are referring to

Accepted as suggested.

361: I suggest "two variables previously analyzed separately..."

Accepted as suggested.

General: There is no statement about the availability of data; it is eLife policy to require all data to be publicly available. Also, it would be helpful to share your code to help understand how you conducted pairwise comparisons, etc.

In the submission it was not mentioned anything about data availability. However, all data and scripts will be uploaded with the VOR if it is required.

Recommendations for the authors:

I found your study interesting and potentially promising. However, there are some fundamental problems with the study design and the hypothesis, including:

<(1) Seasonality simulation - Seasonality is strongly associated with time, so it is unusual to simulate seasonal factors without accounting for time. The actual factors associated with seasonal change in reproductive output may be neither a difference in host blood meal nor temperature and photoperiod. It is therefore, odd to reduce seasonality to a difference in photoperiod and temperature in summer and autumn without even mentioning the time of year when the experiment was carried (except for the mention of February as the time the stock samples were collected from the wild).

The temperature and photoperiod settings are established according to a representative day in both autumn and summer. To determine these settings, we utilized climate data spanning a 3-year period (2020-2022), encompassing the most frequently occurring temperatures and day lengths. The weather conditions remained notably consistent throughout this time frame, which is why the specific year was not mentioned. Moreover, including the year in laboratory experiment details is uncommon, as evident in various papers. This practice can be corroborated by referring to multiple sources (cited in the original manuscript). We mention this in the new version.

(2) Hypothesis - While the hypothesis alludes to the 'reason' for seasonal host shift, the prediction is on the outcome of the interaction between blood meal type and season.

It might be nicer to frame your hypothesis to be consistent with the aim, which is, testing the partial contributions of blood meal type, versus photoperiod and temperature to seasonal change in the reproductive output of Culex quinquefasciatus. A hypothesis like that can be accompanied by alternative predictions according to the expected individual and interactive effects of both factors.

It was rewritten in the revised version to be consistent with our predictions and findings.

Blood meal type, temperature, and photoperiod are all components of seasonality, so the strength of the study is its potential to decouple the effect of blood meal type from that of temperature and photoperiod on the seasonal reproductive output of Culex quinquefasciatus by comparing the two blood meal types under simulated summer and winter conditions. Ideally, this should have been over a natural summer and winter because a natural time difference captures the effect of other seasonal factors other than temperature and photoperiod.

Furthermore, the hypothesis stemmed from field observations, while the study itself was conducted under laboratory conditions using a local population of Culex quinquefasciatus from Argentina. It remains uncertain whether there is supporting evidence for a seasonal shift in host usage in Culex quinquefasciatus from the stock population. Discussing the field observations within the stock population would provide valuable insights.

It was considered in the new version.

Version 2

Abstract

Abstract

Host selection by mosquitoes is a keystone to understanding viral circulation and predicting future infection outbreaks. Culex mosquitoes frequently feed on birds during spring and early summer, shifting into mammals towards late summer and autumn. This host switch may be due to changes in mosquito fitness. The aim of this study was to assess if the interaction effect of blood meal source and seasonality may influence reproductive traits of Culex quinquefasciatus mosquitoes. For this purpose, Cx. quinquefasciatus mosquitoes were reared in simulated summer and autumn conditions and fed on two different hosts, chickens and mice, in a factorial design. Fecundity, fertility, and hatchability during two consecutive gonotrophic cycles were estimated. We found greater fecundity and fertility for mosquitoes fed upon birds than mammals. Fecundity and fertility increased in autumn for chicken-fed mosquitoes, whereas they decreased for mouse-fed mosquitoes. These traits decreased in the second gonotrophic cycle for mouse-fed mosquitoes, whereas they did not vary between cycles for chicken-fed mosquitoes. The effect of hatchability on fertility was rather limited. These results indicate a statistically significant interaction effect of blood meal source and seasonality on fecundity and fertility. However, the pattern was opposite in relation to our hypothesis, suggesting that further studies are needed to confirm and expand our knowledge about mosquito biology and its relationship with seasonal host use shifting.

Introduction

The southern house mosquito, Culex quinquefasciatus Say (Diptera: Culicidae), is a worldwide distributed vector of several pathogens, such as Wuchereria bancrofti Cobbold (Kasili et al., 2009), Dirofilaria immitis Leidy (Labarthe et al., 1998) and avian malaria (Farajollahi et al., 2011). Besides, this mosquito is also one of the main species responsible for the transmission of the arboviruses West Nile virus (WNV) and Saint Louis encephalitis virus (SLEV), among others (Farajollahi et al., 2011). Because of this, this species is of critical concern due to its impact on both public and veterinary health.

Since arboviruses are transmitted by the bite of infected mosquitoes between vertebrate hosts, activity of these pathogens is determined, at least in part, by the blood feeding habits of these vectors. Host preference of Culex mosquitoes is variable and it can be classified in four categories: (1) mammophilic (Muturi et al., 2008), (2) ornithophilic (Jansen et al., 2009), (3) herpetophilic (Janssen et al., 2015) and (4) generalist (Molaei et al., 2007). Additionally, some species may experience a host switch in its feeding habits, from avian to mammal hosts (Hancock & Camp, 2022). These species are of greater concern from an epidemiological standpoint because they can act as bridge vectors in the transmission of some arboviruses, such as SLEV and WNV (Kilpatrick et al., 2005).

Culex quinquefasciatus is generally considered an ornithophilic species in several areas around the globe (Takken & Verhulst, 2013). Despite its ornithophilic nature, this and other Culex species sometimes feed primarily on mammal hosts, including humans, in certain situations. Various studies have pointed out that in autumn there is an increased activity of SLEV and WNV in human populations (Kilpatrick et al., 2006; Spinsanti et al., 2008), suggesting that viral activity spills over to mammals from birds as a result of the mosquito host switch (Edman & Taylor, 1968; Kilpatrick et al., 2006). A few hypotheses have attempted to explain why primarily ornithophilic mosquitoes also feed upon mammals in some contexts. One of the most widely accepted hypothesis states that the shift on feeding behaviour of Culex mosquitoes is a consequence of the autumn migration of American robin (Turdus migratorius), the principal host in the northeastern US, generating an opportunistic change into mammal hosts (Kilpatrick et al., 2006). In South America in general and in Argentina in particular, studies regarding host switch are scarce. Nevertheless, a few limited studies have indicated a seasonal variation throughout the seasons (Stein et al., 2013; Beranek, 2018). Despite that, Argentina has acknowledged seasonal variation in the activity of SLEV in humans, which can be attributed to variations in mosquito host-feeding pattern (Spinsanti et al., 2008). In Southern latitudes, species such as T. migratorius are lacking, hence the “migration hypothesis” is very unlikely for many austral areas, being restricted only for some regions of the US.

A second hypothesis proposed by Burkett-Cadena et al. (2011) suggests that host breeding cycles drive the host shift of mosquitoes. According to this hypothesis, in hosts with parental care, during reproductive seasons there is a greater investment of energy in assuring the offspring survivorship, consequently producing an increase in susceptibility of being bitten by mosquitoes as a result of a decrease in defensive behaviours. This event leads to the detection of peaks of host use during periods of reproductive investment, in summer for birds and autumn for mammals (Burkett-Cadena et al., 2011).

While some other minor hypotheses, unsupported by field data, have been proposed, none of them have been specifically focused on vector biology. Numerous biological factors, including stress, metabolic rate, and the source of a blood meal, as well as environmental variables such as temperature and photoperiod, exert an influence on mosquito physiology that might affect various reproductive traits such as fecundity, development rate or survivorship. This set of factors gives rise to new nutritional requirements that, ultimately, may lead to seasonal variation in host choice by mosquitoes (Ciota et al., 2014; Costanzo et al., 2015; Gervasi et al., 2016; Yan et al. 2017, 2018). In the literature, several studies have evaluated the impact of these biological and environmental variables on mosquito reproduction, as mentioned before. However, due to the complex nature of mosquito biology, there are likely multiple interactions among variables that give rise to new responses that may not be observed individually. Therefore, the aim of the present study was to assess whether exists an interaction effect between the source of the host blood meal and seasonality (in terms of temperature and photoperiod) on three reproductive traits of Culex quinquefasciatus mosquitoes, fecundity, fertility, and hatchability. Our hypothesis states that the interaction between these two variables influence the reproductive outcomes, potentially leading to a seasonal shift in host selection, driven by a reproductive advantage. Considering the reported host preference changes in Cx. quinquefasciatus, in autumn we expect a greater number of eggs (fecundity) and larvae (fertility) in mosquitoes after feeding on a mammal host compared to an avian host, and the opposite relationship in summer.

Materials and methods

Establishment and maintenance of mosquitoes

Egg rafts of Culex quinquefasciatus were collected from a drainage ditch at Universidad Nacional de Córdoba Campus, Córdoba city, in February 2021. Each raft was individually maintained in plastic containers with one liter of distilled water. The hatched larvae were fed with 100 mg of liver powder three times per week until pupation. Pupae were then transferred to plastic emerging cages (21 cm x 12 cm) covered with a tulle-like fabric, containing distilled water but no food. Adults emerging from each raft were identified morphologically (Darsie, 1985) and molecularly (Smith & Fonseca, 2004) to ensure they corresponded to Cx. quinquefasciatus, since Cx. pipiens and its hybrids are also present in Córdoba (Branda et al., 2021). All adults were reared in a cardboard cage of 22.5 liters (28 cm x 36.5 cm) and provided ad libitum with a 10% sugar solution soaked in cotton pads placed on plastic cups. For long-term maintenance of the colony, 24-hour-starved mosquitoes were offered a blood meal from a restrained chicken twice a month. Four days after feeding, a plastic container with distilled water was placed inside the cage to allow engorged females to lay egg rafts. Batches of egg were collected and transferred to plastic containers (30 cm x 25 cm x 7 cm) in a proportion of 3 rafts per container, filled with three liters of distilled water. The hatched larvae were also fed with liver powder at the same proportion described above. Pupae were transferred to emerging cages, and adult mosquitoes were placed in the final cardboard cage. The colony has been maintained for over twenty generations in the Insectary of the Instituto de Virología “Dr. J.M. Vanella” (InViV). Room conditions were controlled at 28°C, with a 12L:12D photoperiod and 70% relative humidity.

Experimental design

Blood source

For experimental trials, avian hosts (live chicks of the species Gallus gallus) and mammalian hosts (live mice of the species Mus musculus, strain C57BL/6) were used to evaluate the effect of blood meal source on fecundity, fertility, and hatchability. Chicks were generously donated by the Bartolucci poultry farm (Córdoba, Argentina), while mice were commercially obtained from the Instituto de Investigación Médica Mercedes y Martín Ferreyra (CONICET - Universidad Nacional de Córdoba). In each trial, 24-hour-starved adult female mosquitoes were provided with a blood meal from restrained chicks or mice. Vertebrate hosts were offered to mosquitoes one hour before the lights were turned off and were kept for 3 hours during 2 consecutive gonotrophic cycles.

The experimental use of animals (G. gallus and M. musculus) was approved by the ethical committee at Facultad de Ciencias Médicas, Universidad Nacional de Córdoba (FCM-UNC) in compliance with the legislation regarding the use of animals for experimental and other scientific purposes (accession code: CE-2022-00518476-UNC-SCT#FCM).

Seasonality

To assess the effect of seasonality (photoperiod + temperature) on fecundity, fertility, and hatchability, a “typical” summer and autumn day from Córdoba city was simulated in an incubator, where mosquitoes were housed throughout the entire experiment. Summer conditions were as follows: T°min = 22°C, T°max = 28°C, photoperiod = 14L:10D. Autumn conditions were characterized by: T°min = 16°C, T°max = 22°C, and photoperiod = 10L:14D. The humidity levels were maintained at 60-70% for both conditions. Data for simulated conditions were obtained from Climate Data website (https://es.climate-data.org/).

Feeding trial design

The interaction effect of blood source (bird or mammal) and seasonality (autumn and summer) was evaluated during two consecutive gonotrophic cycles (I and II). Eight experimental colonies were established using egg rafts collected from the maintenance colony: mammal-autumn-I (MA-I), mammal-summer-I (MS-I), bird-autumn-I (BA-I), bird-summer-I (BS-I), mammal-autumn-II (MA-II), mammal-summer-II (MS-II), bird-autumn-II (BA-II), and bird-summer-II (BS-II).

The experimental colonies were maintained under controlled conditions to ensure that all adults were of the same size (Kauffman et al., 2017). Adult mosquitoes were placed in cardboard cages of 8.3 liters (21 cm x 24 cm) and were provided with ad libitum access to a 10% sugar solution, which was soaked in cotton pads placed on plastic cups.

Feeding trials were conducted at two time points: 5 days post-emergence (first cycle) and 14 days post-emergence (second cycle). Following each blood meal, female mosquitoes were anesthetized using CO2 and classified as fully engorged (VI Sella’s stage), partially fed (I-V Sella’s stage), or unfed (I Sella’s stage), following the classification by Silva Santos et al. (2019). Fully engorged females were counted and separated into a separate cage to complete their gonotrophic cycle. Four days after feeding, a cup containing distilled water was placed inside the cage to facilitate oviposition. After each oviposition cycle, egg rafts were collected, counted, and transferred to a 12-well plate containing 3 mL of distilled water. They were then photographed to subsequently determine the number of eggs per raft. Rafts were maintained in each well until they hatched into L1 larvae, at which point they were preserved with 1 mL of 96% ethanol, and the number of L1 larvae per raft was counted.

Statistical analysis

The reproductive outputs measured for each experimental colony were fecundity, fertility, and hatchability. All analyses were performed using R Studio statistic software, v.4.2.1 (R Core Team, 2022).

Fecundity was defined as the number of eggs per raft and fertility as the number of L1 larvae hatched per raft. To evaluate the effect of blood meal source and seasonality on these two variables, a Generalized Linear Model (GLM; package MASS) with negative binomial error distribution and logarithmic link function was adjusted (Venables & Ripley, 2002). Fecundity and fertility served as the response variables, while the explanatory variables included blood source, seasonality, and gonotrophic cycle, all in a three-way interaction.

Hatchability (= hatching rate) was defined as the ratio of the number of larvae to the number of eggs per raft. A GLM was applied using a quasipoisson error distribution and a logarithmic link function. The response variable was the number of larvae, while the explanatory variables included blood source, seasonality and gonotrophic cycle as interaction effects. Additionally, the logarithm of the number of eggs was incorporated as an offset in the model.

Multiple comparisons among treatments were conducted using the Tukey’s honestly significant difference test (HSD Tukey), incorporating the Kramer (1956) correction for unbalanced data, also known as the Tukey-Kramer method. This approach allows to set the family-wise error rate to 0.05. These contrasts were performed using the multcomp package (Hothorn et al., 2008).

Model validation was assessed using a simulation-based approach to generate randomized quantile residuals to test goodness of fit (Q-Q plot), homoscedasticity, and outliers and influential points (Cook’s distance and Leverage plots). These analyses were conducted with the DHARMa package (Hartig, 2022).

Model accuracy via simulations for the fecundity and fertility models

To assess the accuracy of the results, a simulation approach was conducted using a Monte Carlo algorithm to simulate data and model parameters according to the methodology proposed by Carsey & Harden (2013). For this purpose, parameters including coefficients, treatment means and dispersion parameter (referred to as theta, θ), were extracted from the fecundity and fertility models. Initially, a new dataset consisting of 1000 values (sample size of 125 per treatment) was generated. This was achieved using the rnbinom function from the MASS package, specifying the treatment means (computed from model coefficients) and the theta parameter. Subsequently, a negative binomial GLM was fitted to the simulated data, and the expected means for each of the eight treatments were obtained from the model. This simulation process was repeated 1000 times. The ratio between the number of iterations in which the simulated means fell within the observed 95% confidence interval divided by the total number of simulations was computed as an approximation of the robustness of results.

Results

Effect of blood source and seasonality on fecundity and fertility

Two interaction effects were found to be statistically significant in both the fecundity and fertility models (Tables 1-4). The first interaction effect was found between blood source and seasonality (fecundity: LRT X2 = 25.0, p = 7.72 x 10-7; fertility: LRT X2 = 13.82, p = 0.002), meaning that the effect of blood meal source on the number of eggs and larvae per raft varied with season (Tables 2, 3). In general, mosquitoes that fed on avian blood sources exhibited greater fecundity and fertility than those that fed on mammalian blood source, regardless of the season. When treatments were examined by pairwise comparisons, fecundity and fertility were 13% and 18% greater in autumn compared to summer for bird-fed mosquitoes, while they were 17% and 19% lower for mammal-fed mosquitoes (Fig. 1 A, C, Table 1). The second statistically significant interaction was between blood meal source and the gonotrophic cycle (fecundity: LRT X2 = 9.29, p = 0.0023; fertility: LRT X2 = 14.4, p = 0.00073). This indicated that the effect of blood type on the number of eggs and larvae per raft differed between the first and second feeding trial (Tables 2, 3). For mosquitoes that fed on mammalian blood, fecundity and fertility were 29% lower in the second cycle compared to the first. Although the difference was not statistically significant, fecundity and fertility showed a slight increase in the second cycle for bird-fed mosquitoes. During the second cycle, these outcomes were 41% and 40% higher, respectively, for bird-fed mosquitoes compared to mammal-fed mosquitoes (Fig. 1 B, D, Table 1).

Interaction plots showing the marginal effects of blood meal source-seasonality and blood meal source-gonotrophic cycle on (A, B) fecundity and (C, D) fertility. Predicted mean values and its corresponding confidence intervals are shown. Means sharing same letters are not statistically different.

Effect of blood source and seasonality on hatchability

The overall hatchability reached 86%, with a range spanning from 76% for MA-II to a peak of 96% for MA-I, as outlined in Table 1 and depicted in Figure 2. The model found a statistically significant effect of blood meal source (LRT X2 = 73.06, p = 0.0012) and an interaction effect between seasonality and gonotrophic cycle (LRT X2 = 41.02, p = 0.0015) on hatching rates (Fig. 3, Table 4). Pairwise comparisons revealed a statistically significant 15% decrease in hatchability for mosquitoes from autumn from the first to the second gonotrophic cycle. Additionally, mosquitoes fed on mammals had greater hatchability than those fed on birds. In summer, hatchability during the first cycle was also higher than during the second cycle. These two trends were not found to be statistically significant.

Boxplots showing the effects of both blood meal source and seasonality on hatchability for the first and second gonotrophic cycle.

Interaction plot showing the marginal effects of seasonality-gonotrophic cycle on hatchability.

Model accuracy via simulations for the fecundity and fertility models

When comparing the simulated means for each of the eight treatments (combinations of blood meal source, seasonality, and gonotrophic cycle) with the means calculated from the observed data a high level of accuracy exceeding 90% was observed. Most of the simulated means fitted within the original 95% confidence interval, with an accuracy rate greater than 98%. The greatest mismatching among simulated and original data were found in the treatment MS-I, with an accuracy of 93% and 95% for the fecundity and fertility model, respectively. Accuracy remained high despite variations in sample sizes and data dispersion (Figs. 4, 5).

Simulated means of fecundity (eggs per raft) of each treatment along 1,000 iterations. Abbreviations. B: avian blood; M: mammalian blood; A: autumn; S: summer; I: first gonotrophic cycle; II: second gonotrophic cycle; Ac: model accuracy.

Simulated means of fertility (eggs per raft) of each treatment along 1,000 iterations. Abbreviations same as Figure 4.

Discussion

Our results revealed that there exists an interaction effect of blood meal source and seasonality on reproductive outputs of Cx. quinquefasciatus. Although we expected a higher fecundity and fertility in summer for bird-fed mosquitoes and higher outputs in autumn for mammal-fed mosquitoes, we found the opposite pattern.

Most studies have reported a greater fecundity and fertility in Cx. quinquefasciatus fed on different bird species (Akoh et al.,1993; Richards et al., 2012; Telang & Skinner,2019). This trend is also present in other species of Culex (Shroyer & Silvery, 1972) and Aedes mosquitoes (Harrison et al., 2021). The widely accepted hypothesis that has emerged to explain why avian blood produces a greater fecundity and fertility even in mosquitoes with a known mammal-preferred host choice focuses on the nutritional quality of host blood. Usually, it is addresses that because avian blood possesses nucleated and larger erythrocytes this offers greater quantities of proteins and energy than mammalian blood (Alto et al., 2014). However, this assumption lacks solid evidence due to studies regarding nutritional quality of avian blood and its impact in mosquito fitness are scarce. Actually, some studies have shown that mammalian blood are nutritionally richer than avian blood in terms of total protein content (Pearce 1977, Sant’Anna et al., 2010). Pre-diuresis, a physiological process where red blood cells and proteins are concentrated in the midgut by excreting large amounts of water, has been postulated as a mechanism for explaining the higher fecundity and fertility observed in bird-fed mosquitoes (Erram et al., 2022). Furthermore, avian blood has higher concentrations of isoleucine than mammalian blood, a crucial limiting amino acid necessary for egg production (Greenberg, 1951, Harrison et al., 2021).

The effect of temperature on fitness follows a unimodal relationship with a narrow optimal range. For mosquitoes of genus Culex and Aedes, this optimal band ranges from 20°C to 30°C, with an optimal at 28°C for species such as Cx. quinquefasciatus (Mordecai et al., 2019). Several studies have corroborated this trend, with higher fecundity and fertility towards the optimal of 28°C (Abouzied, 2017; Mogi, 1992), regardless of the blood source (bird or mammal). Regarding photoperiod, the effects are not as evident as temperature since data are somewhat contradictory. While Mogi (1992) found that mid and long day lengths induced greater fecundity, Costanzo et al. (2015) did not find differences of day length on fecundity. Because temperature and photoperiod are always coupled, in general their effects follow the same unimodal trend. Although our experiment was conducted within this optimal range, the highest fecundity and fertility in optimal conditions (summer) was present only in mammal-fed mosquitoes, while higher fitness in bird-fed mosquitoes was detected in autumn. This difference may indicate an interaction of seasonality with blood source.

While our findings may have certain limitations due to moderate sample sizes in some treatments and their reliance solely on laboratory experiments, they verify an interaction effect between blood meal source and seasonality, which had not been previously investigated or documented in any other study. Nevertheless, the precise reasons behind the observed pattern remain inconclusive. The highest levels of fecundity and fertility observed in bird-fed mosquitoes during autumn, when conditions are more challenging, may confer a reproductive advantage, leading to increased offspring production as an adaptive response to these adverse conditions. In contrast, during the more favorable conditions of summer, this pattern may not be as readily discernible. This proposed hypothesis needs further studies to be confirmed.

A second interaction was observed between blood meal source and gonotrophic cycle. We found that fecundity and fertility during the first cycle were similar between mosquitoes fed on chickens and those fed on mammals, and this was also similar with bird-fed mosquitoes from the second gonotrophic cycle, regardless of the season. The only difference we observed was between the fitness of mammal-fed mosquitoes from the second gonotrophic cycle and the earlier mentioned treatments. This pattern is not different from those found in other studies, as the first and second cycles are generally the most productive (Awahmukalah & Brooks, 1985; Christiansen-Jucht et al., 2015). Although we observed a slight increase of fecundity and fertility in bird-fed mosquitoes between cycles, it was statistically non-significant. To our knowledge, only two studies have reported this increasing pattern: an older one by Bennett (1970) and unpublished data from Obholz (personal communication), both conducted with Ae. aegypti. On the contrary, a decrease in fitness between cycles has been reported in other studies for Cx. quinquefasciatus (Richards et al., 2012; Telang & Skinner, 2019) fed upon birds or mammals. This is consistent with our findings regarding mouse-fed mosquitoes; however, because the fecundity and fertility of mouse-fed mosquitoes during the second cycle yielded the fewest egg rafts, this decreasing pattern might also be an effect due to the insufficient sample size.

In addition to assessing fecundity, we also investigated whether hatchability could account for the variations in fertility observed among different treatments. The only statistically significant trend we identified was a decrease in the percentage of hatched eggs from the first to the second gonotrophic cycle, but this was observed only in autumn mosquitoes. This decline in hatching rate may be attributed to a constant exposure to cooling conditions, due to the challenging environmental conditions during this season. This temperature-related effect aligns with findings from other studies (van der Linde et al., 1990). However, although our data suggest that changes in fertility may be linked to variations in hatching rate, it becomes apparent that fecundity has a more pronounced impact on larval production due to the effects of blood meal source and seasonality than hatching rate. Hence, the influence of hatchability on the number of larvae per raft is rather limited.

In summary, our study investigates the critical issue of host shifting patterns in Culex mosquitoes, which holds significant importance due to their role as bridge vectors between bird and human hosts. Our model suggests that changes in photoperiod and environmental temperature induce a different physiological state that requires the mosquito to feed on a blood meal source that results in higher fitness (increased fecundity and fertility). We predicted that mosquitoes fed on mammals in autumn will perform higher fecundity and fertility than those fed on birds in the same season. Unexpectedly, we observed the opposite pattern. Despite this, our study confirmed a significant interaction between blood meal source and seasonality, suggesting a complex interplay of factors affecting mosquito reproduction and its relationship with host use. Due to the observed opposite trend, we can consider three plausible explaining scenarios. Firstly, it is possible that additional factors, such as genetic mechanisms or variations in midgut microbiota composition, play a role in influencing host use patterns. Secondly, host switching might be explained by factors related to vector biology other than fitness. Lastly, it remains a possibility that Argentinian Cx. quinquefasciatus populations do not exhibit the same host-switching behaviour observed in US populations, and the observed interaction effects may be linked to different mosquito behaviours. Further investigations are warranted to corroborate our findings and gain a deeper understanding of the interaction between mosquito feeding pattern and seasonality. Ultimately, these investigations will provide valuable insights into mosquito-borne disease transmission and forecasting its emergence.

eLife Assessment

This useful study provides the first assessment of the potentially interactive effects of seasonality and blood source on mosquito fitness, together in one study. However, the experimental approach is incomplete because it is limited without replication of the experiments and because of the small sample sizes for some groups. The work will be of interest to those studying mosquito biology.

Reviewer #1 (Public Review):

Summary: This study examines the role of host blood meal source, temperature, and photoperiod on the reproductive traits of Cx. quinquefasciatus, an important vector of numerous pathogens of medical importance. The host use pattern of Cx. quinquefasciatus is interesting in that it feeds on birds during spring and shifts to feeding on mammals towards fall. Various hypotheses have been proposed to explain the seasonal shift in host use in this species but have provided limited evidence. This study examines whether the shifting of host classes from birds to mammals towards autumn offers any reproductive advantages to Cx. quinquefasciatus in terms of enhanced fecundity, fertility, and hatchability of the offspring. The authors found no evidence of this, suggesting that alternate mechanisms may drive the seasonal shift in host use in Cx. quinquefasciatus.

Strengths: Host blood meal source, temperature, and photoperiod were all examined together.

Weaknesses: The study was conducted in laboratory conditions with a local population of Cx. quinquefasciatus from Argentina. I'm not sure if there is any evidence for a seasonal shift in the host use pattern in Cx. quinquefasciatus populations from the southern latitudes.

Comments on the revision:

Overall, I am not quite convinced about the possible shift in host use in the Argentinian populations of Cx. quinquefasciatus. The evidence from the papers that the authors cite is not strong enough to derive this conclusion. Therefore, I think that the introduction and discussion parts where they talk about host shift in Cx. quinquefasciatus should be removed completely as it misleads the readers. I suggest limiting the manuscript to talking only about the effects of blood meal source and seasonality on the reproductive outcomes of Cx. quinquefasciatus.

Reviewer #2 (Public Review):

Summary:

Conceptually, this study is interesting and is the first attempt to account for the potentially interactive effects of seasonality and blood source on mosquito fitness, which the authors frame as a possible explanation for previously observed host-switching of Culex quinquefasciatus from birds to mammals in the fall. The authors hypothesize that if changes in fitness by blood source change between seasons, higher fitness on birds in the summer and on mammals in the autumn could drive observed host switching. To test this, the authors fed individuals from a colony of Cx. quinquefasciatus on chickens (bird model) and mice (mammal model) and subjected each of these two groups to two different environmental conditions reflecting the high and low temperatures and photoperiod experienced in summer and autumn in Córdoba, Argentina (aka seasonality). They measured fecundity, fertility, and hatchability over two gonotrophic cycles. The authors then used a generalized linear model to evaluate the impact of host species, seasonality, and gonotrophic cycle on fecundity, fertility, and hatchability. The authors were trying to test their hypothesis by determining whether there was an interactive effect of season and host species on mosquito fitness. This is an interesting hypothesis; if it had been supported, it would provide support for a new mechanism driving host switching. While the authors did report an interactive impact of seasonality and host species, the directionality of the effect was the opposite from that hypothesized. The authors have done a very good job of addressing many of the reviewer concerns, with several exception that continue to cause concern about the conclusions of the study.

Strengths:

(1) Using a combination of laboratory feedings and incubators to simulate seasonal environmental conditions is a good, controlled way to assess the potentially interactive impact of host species and seasonality on the fitness of Culex quinquefasciatus in the lab.

(2) The driving hypothesis is an interesting and creative way to think about a potential driver of host switching observed in the field.

(3) The manuscript has become a lot clearer and easier to read with the revisions - thank you to the authors for working hard to make many of the suggested changes.

Weaknesses:

(1) The authors have decided not to follow the suggestion of conducting experimental replicates of the study. This is understandable given the significant investment of resources and time necessary, however, it leaves the study lacking support. Experimental replication is an important feature of a strong study and helps to provide confidence that the observed patterns are real and replicable. Without replication, I continue to lack confidence in the conclusions of the study.

(2) The authors have included some additional discussion about the counterintuitive nature of their results, but the paragraph discussing this in the discussion was confusing. I believe that this should be revised. This is a key point of the paper and needs to be clear to the reader.

(3) There should be more discussion of the host switching observed in the two studies conducted in Argentina referenced by the authors. Since host switching is the foundation for the hypothesis tested in this paper, it is important to fully explain what is currently known in Argentina.

(4) In some cases, the explanations of referenced papers are not entirely accurate. For example, when referencing Erram et al 2022, I think the authors misrepresented the paper's discussion regarding pre-diuresis- Erram et al. are suggesting that pre-diuresis might be the mechanism by which C. furens compensates for the lower nutritional value of avian blood, leading to no significant difference between avian/mammal blood on fecundity/fertility (rather than leading to higher fecundity on birds, as stated in this manuscript). The study performed by Erram et al. also didn't prove this phenomenon, they just suggest it as a possible mechanism to explain their results, so that should be made clear when referencing the paper.

(5) In some cases, the conclusions continue to be too strongly worded for the evidence available. For example, lines 322-324: I don't think the data is sufficient to conclude that a different physiological state is induced, nor that they are required to feed on a blood source that results in higher fitness.

(6) There is limited mention of the caveat that this experiment performed with simulated seasonality that does not perfectly replicate seasonality in the field. I think this caveat should be discussed in the discussion (e.g. that humidity is held constant).

Author response:

The following is the authors’ response to the previous reviews.

We have carefully addressed all the reviewers' suggestions, and detailed responses are provided at the end of this letter. In summary:

• We conducted two additional replicates of the study to obtain more robust and reliable data.

• The Introduction has been revised for greater clarity and conciseness.

• The Results section was shortened and reorganized to highlight the key findings more effectively.

• The Discussion was modified according to the reviewers' suggestions, with a focus on reorganization and conciseness.

We hope you find this revised version of the manuscript satisfactory.

Reviewer #1 (Public Review):

Summary:

This study examines the role of host blood meal source, temperature, and photoperiod on the reproductive traits of Cx. quinquefasciatus, an important vector of numerous pathogens of medical importance. The host use pattern of Cx. quinquefasciatus is interesting in that it feeds on birds during spring and shifts to feeding on mammals towards fall. Various hypotheses have been proposed to explain the seasonal shift in host use in this species but have provided limited evidence. This study examines whether the shifting of host classes from birds to mammals towards autumn offers any reproductive advantages to Cx. quinquefasciatus in terms of enhanced fecundity, fertility, and hatchability of the offspring. The authors found no evidence of this, suggesting that alternate mechanisms may drive the seasonal shift in host use in Cx. quinquefasciatus.

Strengths:

Host blood meal source, temperature, and photoperiod were all examined together.

Weaknesses:

The study was conducted in laboratory conditions with a local population of Cx. quinquefasciatus from Argentina. I'm not sure if there is any evidence for a seasonal shift in the host use pattern in Cx. quinquefasciatus populations from the southern latitudes.

Comments on the revision:

Overall, I am not quite convinced about the possible shift in host use in the Argentinian populations of Cx. quinquefasciatus. The evidence from the papers that the authors cite is not strong enough to derive this conclusion. Therefore, I think that the introduction and discussion parts where they talk about host shift in Cx. quinquefasciatus should be removed completely as it misleads the readers. I suggest limiting the manuscript to talking only about the effects of blood meal source and seasonality on the reproductive outcomes of Cx. quinquefasciatus.

As mentioned in the previous revision, we agree on the reviewer observation about the lack of evidence on seasonal shift in the host use pattern in Cx. quinquefasciatus populations from Argentina. We include this topic in the discussion.

Additionally, we also added a paragraph in the discussion section to include the limitations of our study and conclusions. One of them is the fact that our results are based on controlled conditions experiments. Future studies are needed to elucidate if the same trend is found in the field.

Reviewer #1 (Recommendations for the authors):

Abstract

Line 73: shift in feeding behavior

Accepted as suggested.

Discussion

Line 258: addressed that Accepted as suggested.

Line 263: blood is nutritionally richer

Accepted as suggested.

Reviewer #2 (Public Review):

Summary:

Conceptually, this study is interesting and is the first attempt to account for the potentially interactive effects of seasonality and blood source on mosquito fitness, which the authors frame as a possible explanation for previously observed host-switching of Culex quinquefasciatus from birds to mammals in the fall. The authors hypothesize that if changes in fitness by blood source change between seasons, higher fitness on birds in the summer and on mammals in the autumn could drive observed host switching. To test this, the authors fed individuals from a colony of Cx. quinquefasciatus on chickens (bird model) and mice (mammal model) and subjected each of these two groups to two different environmental conditions reflecting the high and low temperatures and photoperiod experienced in summer and autumn in Córdoba, Argentina (aka seasonality). They measured fecundity, fertility, and hatchability over two gonotrophic cycles. The authors then used a generalized linear model to evaluate the impact of host species, seasonality, and gonotrophic cycle on fecundity, fertility, and hatchability. The authors were trying to test their hypothesis by determining whether there was an interactive effect of season and host species on mosquito fitness. This is an interesting hypothesis; if it had been supported, it would provide support for a new mechanism driving host switching. While the authors did report an interactive impact of seasonality and host species, the directionality of the effect was the opposite from that hypothesized. The authors have done a very good job of addressing many of the reviewer concerns, with several exception that continue to cause concern about the conclusions of the study.

Strengths:

(1) Using a combination of laboratory feedings and incubators to simulate seasonal environmental conditions is a good, controlled way to assess the potentially interactive impact of host species and seasonality on the fitness of Culex quinquefasciatus in the lab.

(2) The driving hypothesis is an interesting and creative way to think about a potential driver of host switching observed in the field.

(3) The manuscript has become a lot clearer and easier to read with the revisions - thank you to the authors for working hard to make many of the suggested changes.

Weaknesses:

(1) The authors have decided not to follow the suggestion of conducting experimental replicates of the study. This is understandable given the significant investment of resources and time necessary, however, it leaves the study lacking support. Experimental replication is an important feature of a strong study and helps to provide confidence that the observed patterns are real and replicable. Without replication, I continue to lack confidence in the conclusions of the study.

We included replicates as suggested.

(2) The authors have included some additional discussion about the counterintuitive nature of their results, but the paragraph discussing this in the discussion was confusing. I believe that this should be revised. This is a key point of the paper and needs to be clear to the reader.

Revised as suggested.

(3) There should be more discussion of the host switching observed in the two studies conducted in Argentina referenced by the authors. Since host switching is the foundation for the hypothesis tested in this paper, it is important to fully explain what is currently known in Argentina.

Accepted as suggested.

(4) In some cases, the explanations of referenced papers are not entirely accurate. For example, when referencing Erram et al 2022, I think the authors misrepresented the paper's discussion regarding pre-diuresis- Erram et al. are suggesting that pre-diuresis might be the mechanism by which C. furens compensates for the lower nutritional value of avian blood, leading to no significant difference between avian/mammal blood on fecundity/fertility (rather than leading to higher fecundity on birds, as stated in this manuscript). The study performed by Erram et al. also didn't prove this phenomenon, they just suggest it as a possible mechanism to explain their results, so that should be made clear when referencing the paper.

Changed as suggested.

(5) In some cases, the conclusions continue to be too strongly worded for the evidence available. For example, lines 322-324: I don't think the data is sufficient to conclude that a different physiological state is induced, nor that they are required to feed on a blood source that results in higher fitness.

Redaction was modified as suggested to tight our discussion with results.

(6) There is limited mention of the caveat that this experiment performed with simulated seasonality that does not perfectly replicate seasonality in the field. I think this caveat should be discussed in the discussion (e.g. that humidity is held constant).

This topic is now included in the discussion as suggested.

Reviewer #2 (Recommendations for the authors):

59-60: These terms should end with -phagic instead of -philic. These papers study blood feeding patterns, not preference. I understand that the Janssen papers calls it "mammalophilic" in their title, but this was an incorrect use of the term in their paper. There are some review papers that explain the difference in this terminology if it's helpful.

Accepted as suggested.

73: edit to "in" feeding behavior

Accepted as suggested.

77-78: Given that the premise of your study is based on the phenomenon of host switching, I suggest that you expand your discussion of these two papers. What did they observe? Which hosts did they switch from / to and how dramatic was the shift?

Accepted as suggested.

79: replace acknowledged with experienced

Accepted as suggested.

79-80: the way that this is written is misleading. It suggests that Spinsanti showed that seasonal variation in SLEV could be attributed to a host shift, which isn't true. This citation should come before the comma and then you should use more cautious language in the second half. E.g which MIGHT be possible to attribute to ....

Accepted as suggested.

80-82: this is not convincing. Even if the Robin isn't in Argentina, Argentina does have migrating birds, so couldn't this be the case for other species of birds? Do any of the birds observed in previous blood meal analyses in Argentina migrate? If so, couldn't this hypothesis indeed play a role?

A paragraph about this topic was added to the discussion as suggested.

90: hypotheses for what? The fall peak in cases? Or host switching?

Changed to be clearer.

98: where was this mentioned before? I think "as mentioned before" can be removed.

Accepted as suggested.

101: edit to "whether an interaction effect exists"

Accepted as suggested.

104: edit to "We hypothesize that..."

Accepted as suggested.

106: reported host USE changes, not host PREFERENCE changes, right?

All the terminology was change to host pattern and not preference to avoid confusion.

200: Briefly reading Carsey and Harden, it looks like the methodology was developed for social science. Is there anything you can cite to show this applied to other types of data? If not, I think this requires more explanation in your MS.

This was removed as replicates were included.

237-239: I think it is best not to make a definitive statement about greater/higher if it isn't statistically significant; I suggest modifying the sentences to state that the differences you are listing were not significantly different up front rather than at the end, otherwise if people aren't reading carefully, they may get the wrong impression.

Accepted as suggested.

245: you only use the term MS-I once before and I forgot what it meant since it wasn't repeated, so I had to search back through with command-F. I suggest writing this out rather than using the acronym.

Accepted as suggested.

249: edit to: "an interaction exists between the effect of..."

Accepted as suggested.

253-254: greater compared to what?

Change for clearness. 258-260: edit for grammar

Accepted as suggested.

260-262: edit for grammar; e.g. "However, this assumption lacks solid evidence; there is a scarcity of studies regarding nutritional quality of avian blood and its impact on mosquito fitness."

Accepted as suggested.

263: edit: blood is nutritionally...

Accepted as suggested.

264-267: This doesn't sound like an accurate interpretation of what the paper suggests regarding pre-diuresis in their discussion - they are suggesting that pre-diuresis might be the mechanism by which C. furens compensates for the lower nutritional value of avian blood, leading to no significant difference between avian/mammal blood on fecundity/fertility. They also don't show this, they just suggest it as a possible mechanism to explain their results.

This topic was removed given the restructuring of discussion.

253-269: You should tie this paragraph back to your results to explicitly compare/contrast your findings with the previous literature.

Accepted as suggested.

270-282: This paragraph would be a good place to explain the caveat of working in the laboratory - for example, humidity was the same across the two seasons which I'm guessing isn't the case in the field in Argentina. You can discuss what aspects of laboratory season simulation do not accurately replicate field conditions and how this can impact your findings. You said in your response to the reviewers that you weren't interested in measuring other variables (which is fair, and not expected!), but the beauty of the discussion section is to be able to think about how your experimental design might impact your results - one possibility is that your season simulation may not have produced the results produced by true seasonal shifts.

Accepted as suggested.

279-281: You say your experiment was conducted within the optimal range, which would suggest that both summer and autumn were within that range, but then you only talk about summer as optimal in the following sentence.

Changed for clearness.

281-282: You should clarify this sentence - state what the interaction has an effect on.

Accepted as suggested.

283-291: I appreciate that your discussion now acknowledges the small sample size and the questions that remain unanswered due to the results being opposite to that of the hypothesis, but this paragraph lacks some details and in places doesn't make sense.

I think you need to emphasize which groups had small sample size and which conclusions that might impact. I also think you need to explain why the sample size was substantially smaller for some groups (e.g. did they refuse to feed on the mouse in the autumn?). I appreciate that sample sizes are hard to keep high across many groups and two gonotrophic periods, but unfortunately, that is why fitness experiments are so hard to do and by their nature, take a long time. I understand that other papers have even lower sample size, but I was not asked to review those papers and would have had the same critique of them. I don't believe that creating simulated data via a Monte Carlo approach can make up for generating real data. As I understand it from your explanation, you are parametrizing the Monte Carlo simulations with your original data, which was small to begin with for autumn mouse. Using this simulation doesn't seem like a satisfactory replacement for an experimental replicate in my opinion. I maintain that at least a second replicate is necessary to see whether the patterns that you have observed hold.

The performing of a power analysis and addition of more replicates tried to solve the issue of sample size. More about this critic is added in the discussion. The simulation approach was totally removed.

Regarding the directionality of the interaction effect, I think this warrants more discussion. Lines 287-291 don't make sense to me. You suggest that feeding on birds in the autumn may confer a reproductive advantage when conditions are more challenging. But then why wouldn't they preferentially feed on birds in the autumn, rather than mammals? I suggest rewriting this paragraph to make it clearer.

Accepted as suggested.

297: earlier mentioned treatments? Do you mean compared to the first gonotrophic cycle? This isn't clear.

Changed for clearness.

302-303: Did you clarify whether you are allowed to reference unpublished data in eLife?

This was removed to follow the guidelines of eLife.

316-317: "it becomes apparent" sounds awkward, I suggest rewording and also explaining how this conclusion was made.

Accepted as suggested.

322-324: I think that this statement is too strongly worded. I don't think your data is sufficient to conclude that a different physiological state is induced, nor that they are required to feed on a blood source that results in higher fitness. Please modify this and make your conclusions more cautious and closely linked to what you actually demonstrated.

Accepted as suggested.

325: change will perform to would have

Accepted as suggested.

326: add to the sentence: "and vice versa in the summer"

Accepted as suggested.

330: possible explanations, not explaining scenarios.

Accepted as suggested.

517: I think you should repeat the abbreviation definitions in the caption to make it easier for readers, otherwise they have to flip back and forth which can be difficult depending on formatting.

Accepted as suggested.

In general, I think that your captions need more information. I think the best captions explain the figure relatively thoroughly such that the reader can look at the figure and caption and understand without reading the paper in depth. (e.g. the statistical test used).

Data availability: The eLife author instructions do say that data must be made available, so there should be a statement on data availability in your MS. I also suggest you make the code available.

Accepted as suggested.

Version 3

Abstract

Abstract

Host selection by mosquitoes is a keystone to understanding viral circulation and predicting future infection outbreaks. Culex mosquitoes frequently feed on birds during spring and early summer, shifting into mammals towards late summer and autumn. This host switch may be due to changes in mosquito fitness. The aim of this study was to assess if the interaction effect of blood meal source and seasonality may influence reproductive traits of Culex quinquefasciatus mosquitoes. For this purpose, Cx. quinquefasciatus mosquitoes were reared in simulated summer and autumn conditions and fed on two different hosts, chickens and mice, in a factorial design. Fecundity, fertility, and hatchability during two consecutive gonotrophic cycles were estimated. We found greater fecundity and fertility for mosquitoes fed upon birds than mammals. Fecundity and fertility did not vary between seasons for chicken-fed mosquitoes, whereas in autumn they decreased for mouse-fed mosquitoes. These traits decreased in the second gonotrophic cycle for mouse-fed mosquitoes, whereas they did not vary between cycles for chicken-fed mosquitoes. There was no statistically significant variation of hatchability among treatments. These results indicate a statistically significant interaction effect of blood meal source and seasonality on fecundity and fertility. However, the pattern was opposite in relation to our hypothesis, suggesting that further studies are needed to confirm and expand our knowledge about mosquito biology and its relationship with seasonal host use shifting.

Introduction

The southern house mosquito, Culex quinquefasciatus Say (Diptera: Culicidae), is a worldwide distributed vector of numerous pathogens, including Wuchereria bancrofti Cobbold (Kasili et al., 2009), Dirofilaria immitis Leidy (Labarthe et al., 1998) and avian malaria (Farajollahi et al., 2011). Besides, this mosquito is also one of the main species responsible for the transmission of the arboviruses West Nile (WNV) and St. Louis encephalitis (SLEV) viruses (Farajollahi et al., 2011). Because this, this species is of critical concern due to its impact on both public and veterinary health.

Since arboviruses are transmitted by the bite of infected mosquitoes between vertebrate hosts, activity of these pathogens is determined, at least in part, by the blood feeding habits of these vectors. According to its host feeding pattern range, Culex mosquitoes can be regarded as specialists or generalists. Specialist species include mosquitoes feeding primarily on mammals (Muturi et al., 2008), birds (Jansen et al., 2009), reptiles and amphibians (Janssen et al., 2015). Generalists are mosquitoes fed on the most abundant available hosts (Molaei et al., 2007). Additionally, some species could experience a seasonal shift in its feeding habits, from avian to mammal hosts (Edman & Taylor, 1968; Hancock & Camp, 2022), being of greater concern from an epidemiological perspective given that they can act as bridge vectors in the transmission of some arboviruses, such as SLEV and WNV (Kilpatrick et al., 2006).

West Nile virus (WNV) and St. Louis encephalitis virus (SLEV) are zoonotic pathogens maintained in nature through an enzootic network involving transmission among viremic birds and ornithophagic mosquitoes (Kilpatrick et al., 2006). However, during late summer and early autumn, these viruses increase their activity in humans and other mammals, which has led to repeated epidemics in the US (Kilpatrick et al., 2006) and South America (Spinsanti et al., 2008). Likewise, several Culex species, including Cx. quinquefasciatus, have exhibited a seasonal shift in their feeding behaviour, feeding primarily on birds in spring and early summer and increasing the proportion of mammal blood meals in late summer and autumn (Thiemann et al., 2011). This seasonal shift suggests that viral activity might spill over from birds to mammals as a result of the mosquito host switch (Edman & Taylor, 1968; Kilpatrick et al., 2006).

Two main contrasting hypotheses have been proposed to explain why primarily ornithophagic mosquitoes increase their feeding on mammals later in the season. The migration hypothesis, proposed by Kilpatrick et al. (2006) for Cx. pipiens, suggests that this shift in feeding pattern is related to the abundance of the preferred avian host, the American robin (Turdus migratorius L.). This hypothesis is based on their findings that as robin densities decrease, the proportion of mammalian blood meals in Cx. pipiens populations increase. However, this hypothesis has several limitations: (i) multiple studies have shown that American robins are not always the preferred host for all Cx. pipiens populations (Patrican et al., 2007; Apperson et al., 2002); (ii) in some areas (e.g., Tennessee) with documented seasonal host shifts, American robins are present year-round (Savage et al., 2007); and (iii) this explanation applies only to mosquitoes whose primary host is the American robin, a species found only in the Northern Hemisphere, or another migratory bird. Therefore, it is not applicable to southern regions, such as Argentina, where SLEV seasonality is documented (Spinsanti et al., 2008) but American robins are absent. Because this, it seems that the hypothesis proposed by Kilpatrick et al. (2006) might be habitat dependent (Caranci, 2010). On the other hand, the defensive behaviour hypothesis proposed by Burkett-Cadena et al. (2011), suggests that host breeding cycles drive the host shift of mosquitoes. According to this hypothesis, in hosts with parental care, during reproductive seasons there is a greater investment of energy in assuring the offspring survivorship, consequently producing an increase in susceptibility of being bitten by mosquitoes as a result of a decrease in defensive behaviours. This event leads to the detection of peaks of host use during periods of reproductive investment, in summer for birds and autumn for mammals (Burkett-Cadena et al., 2011). While this hypothesis was proposed as an alternative to address the limitations of the migration hypothesis, it also has drawbacks. It is based on a limited range of host species and assumes a distinct seasonal phenology, which is not present for all mosquito hosts. Because this, this hypothesis seems to be host dependent.

Explanations for the seasonal shift in mosquito host feeding focused on vector biology are largely lacking in the literature. Numerous biological factors, such as stress, metabolic rate, and blood meal source, along with environmental variables like temperature and photoperiod, influence mosquito physiology and can affect reproductive traits such as fecundity, development rate, and survivorship. These factors can give rise to new nutritional requirements that may ultimately lead to seasonal variations in host selection by mosquitoes (Ciota et al., 2014; Costanzo et al., 2015; Gervasi et al., 2016; Yan et al., 2017, 2018). While several studies have examined the effects of these biological and environmental variables on mosquito reproduction, the complex nature of mosquito biology suggests that multiple interactions among these variables may produce novel responses that are not apparent when considered individually. The aim of this study was to assess whether there is an interaction between the source of the host blood meal and seasonality (defined by temperature and photoperiod) on three reproductive traits of Culex quinquefasciatus mosquitoes: fecundity, fertility, and hatchability. We hypothesize that the interaction between these two variables influence the reproductive outcomes, potentially leading to a seasonal shift in host selection, driven by a reproductive advantage. Given the reported seasonal changes in host use by Cx. quinquefasciatus, in autumn we expect a greater number of eggs (fecundity) and larvae (fertility) in mosquitoes after feeding on a mammal host compared to an avian host, and the opposite trend in summer.

Materials and methods

Establishment and maintenance of mosquitoes

Egg rafts of Culex quinquefasciatus were collected from a drainage ditch at Universidad Nacional de Córdoba Campus, Córdoba city, in February 2021. Each raft was individually maintained in plastic containers with one liter of distilled water. The hatched larvae were fed with 100 mg of liver powder three times per week until pupation. Pupae were then transferred to plastic emerging cages (21 cm x 12 cm) covered with a tulle-like fabric, containing distilled water but no food. Adults emerging from each raft were identified morphologically (Darsie, 1985) and molecularly (Smith & Fonseca, 2004) to ensure they corresponded to Cx. quinquefasciatus, since Cx. pipiens and its hybrids coexist sympatrically in Córdoba city (Branda et al., 2021). All adults were reared in a cardboard cage of 22.5 liters (28 cm x 36.5 cm) and provided ad libitum with a 10% sugar solution soaked in cotton pads placed on plastic cups. For long-term maintenance of the colony, 24-hour-starved mosquitoes were offered a blood meal from a restrained chicken twice a month. Four days after feeding, a plastic container with distilled water was placed inside the cage to allow engorged females to lay egg rafts. Batches of egg were collected and transferred to plastic containers (30 cm x 25 cm x 7 cm) in a proportion of three rafts per container, filled with three liters of distilled water. The hatched larvae were also fed with liver powder at the same proportion described above. Pupae were transferred to emerging cages, and adult mosquitoes were placed in the final cardboard cage. The colony has been maintained for several generations in the Insectary of the Instituto de Virología “Dr. J.M. Vanella” (InViV). Room conditions were controlled at 28 ± 1 °C, with a 12L:12D photoperiod and 70% relative humidity.

Experimental design

Blood source

For experimental trials, avian hosts (live chicks of the species Gallus gallus) and mammalian hosts (live mice of the species Mus musculus, strain C57BL/6) were used to evaluate the effect of blood meal source on fecundity, fertility, and hatchability. Chicks were generously donated by the Bartolucci poultry farm (Córdoba, Argentina), while mice were commercially obtained from the Instituto de Investigación Médica Mercedes y Martín Ferreyra (CONICET - Universidad Nacional de Córdoba). In each trial, 24-hour-starved adult female mosquitoes were provided with a blood meal from restrained chicks or mice. Vertebrate hosts were offered to mosquitoes one hour before the lights were turned off and were kept for 3 hours during 2 consecutive gonotrophic cycles.

The experimental use of animals (mosquitoes and vertebrates) was approved by the ethical committee at Facultad de Ciencias Médicas, Universidad Nacional de Córdoba (FCM-UNC) in compliance with the legislation regarding the use of animals for experimental and other scientific purposes (accession code: CE-2022-00518476-UNC-SCT#FCM).

Seasonality

To assess the effect of seasonality (photoperiod + temperature) on fecundity, fertility, and hatchability, a “typical” summer and autumn day from Córdoba city was simulated in an incubator where mosquitoes were housed throughout the entire experiment. Summer conditions were as follows: T°min = 22°C, T°max = 28°C, photoperiod = 14L:10D. Autumn conditions were characterized by: T°min = 16°C, T°max = 22°C, and photoperiod = 10L:14D. The humidity levels were maintained at 60-70% for both conditions. Data for simulated conditions were obtained from Climate Data website (https://es.climate-data.org/).

Feeding trial design

The interaction effect of blood source (chicken and mouse) and seasonality (autumn and summer) was evaluated during two consecutive gonotrophic cycles (I and II). This combination produced eight colonies that were established using egg rafts collected from the maintenance colony: mouse:autumn-I, mouse:autumn-II, chicken:autumn-I, chicken:autumn-II, mouse:summer-I, mouse:summer-II, chicken:summer-I, chicken:summer-II. This set of 8 colonies was repeated in triplicate (1-3), generating a total of 24 treatments.

The experimental colonies were maintained under controlled conditions to ensure that all adults were of the same size (Kauffman et al., 2017). Adult mosquitoes were placed in cardboard cages of 8.3 liters (21 cm x 24 cm) and were provided with ad libitum access to a 10% sugar solution, which was soaked in cotton pads placed on plastic cups.

Feeding trials were conducted at two time points: 5 days post-emergence (first cycle) and 14 days post-emergence (second cycle). Following each blood meal, female mosquitoes were anesthetized using CO2 and classified as fully engorged (VI Sella’s stage), partially fed (I-V Sella’s stage), or unfed (I Sella’s stage), following the classification by Silva Santos et al. (2019). Fully engorged females were counted and separated into a separate cage to complete their gonotrophic cycle. Four days after feeding, a cup containing distilled water was placed inside the cage to facilitate oviposition. After each oviposition cycle, egg rafts were collected, counted, and transferred to a 12-well plate containing 4 mL of distilled water. They were then photographed to subsequently determine the number of eggs per raft. Rafts were maintained in each well until they hatched into L1 larvae, at which point they were preserved with 2 mL of 96% ethanol, and the number of L1 larvae per raft was counted.

Statistical analysis

The reproductive outputs measured for each experimental colony were fecundity, fertility, and hatchability. All analyses were performed using R Studio statistic software, v.4.2.1 (R Core Team, 2022).

Fecundity was defined as the number of eggs per raft and fertility as the number of L1 larvae hatched per raft. To evaluate the effect of blood meal source and seasonality on these two variables, a Generalized Linear Mixed Model (GLMM; lme4 package) with negative binomial error distribution and logarithmic link function was adjusted (Bates et al., 2015). Fecundity and fertility served as the response variables, while the fixed explanatory variables included blood source, seasonality, and gonotrophic cycle, all in a three-way interaction. Treatments (n = 24) were set as a random intercept.

Hatchability (= hatching rate) was defined as the ratio of the number of larvae to the number of eggs per raft. A Generalized Linear Model (GLM; MASS package) was applied using a quasipoisson error distribution and a logarithmic link function (Venables et al., 2002). The response variable was the number of larvae, while the explanatory variables included blood source, seasonality, gonotrophic cycle and replicate as interaction effects. Additionally, the logarithm of the number of eggs was incorporated as an offset in the model.

Multiple comparisons among treatments were conducted using Tukey’s honestly significant difference test (HSD Tukey), incorporating the Kramer (1956) correction for unbalanced data, also known as the Tukey-Kramer method. This approach allows setting the family-wise error rate to 0.05. These comparisons were performed using the emmeans package (Lenth, 2024).

Model validation was assessed using a simulation-based approach to generate randomized quantile residuals to test goodness of fit (Q-Q plot), homoscedasticity, and outliers and influential points (Cook’s distance and Leverage plots). These analyses were conducted with the DHARMa package (Hartig, 2022).

Additionally, the statistical power of the models was assessed using the mixedpower package (Kumle et al., 2021), specifically designed to evaluate the power of Generalized Linear Mixed Models (GLMMs). The analysis involved simulating 1,000 iterations with a sample size ranging from 30 to 150 observations per group. The power analysis aimed to achieve a power threshold of 0.8, indicating an 80% chance of detecting true effects if they exist.

Results

A total of 1,162 egg rafts were obtained for analysis, distributed across the eight treatments as follows: chicken:autumn-I: 260, chicken:autumn-II: 174, chicken:summer-I: 162, chicken:summer-II: 88, mouse:autumn-I: 146, mouse:autumn-II: 96, mouse:summer-I: 167, and mouse:summer-II: 69. The full count of egg rafts for each replicate (24 treatments), along with the means for fecundity, fertility, and hatchability, are summarized in Table 1. This sample size fulfilled the statistical power of 0.8 (Figure SM1).

In both models—fecundity and fertility—the interaction between blood source and seasonality was statistically significant (fecundity: LRT X² = 5.69, p < 0.05; fertility: LRT X² = 4.37, p < 0.05) (Tables 2,3; Figures 1,2A). Mosquitoes that fed on chicken blood had the highest fecundity (summer: 136.4 eggs/raft, autumn: 141.5 eggs/raft) and fertility (summer: 123.9 larvae/raft, autumn: 126 larvae/raft) in both seasons. In summer, fecundity and fertility were 7% and 11% higher, respectively, in chicken-fed mosquitoes compared to mouse-fed mosquitoes. In autumn, these differences increased to 46% for fecundity and 52% for fertility in favor of chicken-fed mosquitoes. No significant seasonal variation was observed in fecundity and fertility for chicken-fed mosquitoes, but both measures were statistically significant lower by 24% and 25%, respectively, in autumn compared to summer for mouse-fed mosquitoes.

Interaction plot showing the effect of blood meal source (chicken or mouse) and seasonality (autumn or summer) on fecundity (eggs/raft). Points represent the predicted marginal mean values and the bars corresponding to their confidence intervals. Means sharing the same letters are not statistically different (p > 0.05).

Interaction plots showing the effect of blood meal source (chicken and mouse) and seasonality (autumn and summer) (A) or blood meal source and gonotrophic cycle (first or second) (B) on fertility (eggs/raft) of Culex quinquefasciatus. Points represent the predicted marginal mean values and the bars corresponding to their confidence intervals. Means sharing the same letters are not statistically different (p > 0.05).

In the fertility model, there was also a significant interaction between blood source and gonotrophic cycle (LRT X² = 3.98, p < 0.05), indicating that the impact of blood type on larvae production differed between the first and second cycles (Table 3, Figure 2B). Chicken-fed mosquitoes showed no difference in fertility between cycles, while those fed on mouse blood had 25% lower fertility in the second cycle compared to the first. During the second cycle, fertility was 50% higher in chicken-fed mosquitoes compared to mouse-fed mosquitoes.

Figure 3 presents the predicted mean values of fecundity and fertility for the 24 treatments, with no differences observed across the three replicates. The variance of random intercepts among treatments was approximately 0.006 for fecundity and 0.007 for fertility, accounting for 6% and 4% of the total variance in the respective models.

Dotplots showing the variation of predicted fecundity (eggs/raft) and fertility (larvae/raft) means along the 24 treatments, combination of blood source (chicken or mouse), seasonality (autumn or summer), gonotrophic cycle (first or second) and replicate (1 to 3).

There was no statistical difference among treatments and replicates in terms of hatchability, given that no statistical effects of blood source, seasonality or gonotrophic cycle was observed (Table 4). The mean hatchability for all treatments was 0.89, with a range between 0.84 for the mouse:autumn-II (replicate 1) colony and 0.96 for mouse:autumn-I (replicate 1) (Table 1, Figure 4).

Boxplots showing the mean hatchability for the 24 treatments, combining replicate (1-3), blood meal source (chicken and mouse), seasonality (summer and autumn) and gonotrophic cycle (first and second), of Culex quinquefasciatus.

Discussion

Our results show there is an interaction effect between blood meal source and seasonality on reproductive outputs of Cx. quinquefasciatus. While our model hypothesized that fecundity and fertility would be higher in bird-fed mosquitoes than in mammal-fed mosquitoes during summer, and the reverse would occur in autumn, our findings revealed the opposite pattern. Regardless the season, fecundity and fertility in Culex quinquefasciatus were higher in bird-fed mosquitoes compared to those fed on mammals. This pattern aligns with those findings from previous studies (Akoh et al., 1993; Richards et al., 2012; Telang & Skinner, 2019). This trend is not unique to Cx. quinquefasciatus but has also been seen in other Culex species (Shroyer & Silvery, 1972) and Aedes mosquitoes (Harrison et al., 2021). The commonly accepted explanation for this pattern suggests that the nutritional quality of avian blood, with its nucleated and larger erythrocytes, provides greater protein and energy, thereby enhancing these reproductive outputs (Alto et al., 2014). However, our findings offer a mixed perspective by highlighting the interaction between blood meal source and seasonality that further might influence fecundity and fertility.

The significant interaction observed between blood meal source and seasonality revealed that mammal-fed mosquitoes had lower fecundity and fertility in autumn compared to summer, while bird-fed mosquitoes maintained similar reproductive outputs across both seasons. This finding diverges from the expectation of consistently higher fecundity and fertility in bird-fed mosquitoes and consistently lower fecundity and fertility in mammal-fed mosquitoes. This suggests that environmental factors such as temperature and photoperiod, which have been shown to affect mosquito fitness (Mordecai et al., 2019; Abouzied, 2017; Mogi, 1992), may interact with blood meal source in complex ways. Although the effects of temperature and photoperiod are generally coupled and follow a unimodal trend with an optimal peak at 28°C (Mordecai et al., 2019), our results indicate that the interaction between seasonality and blood source might play a critical role in determining reproductive success.

Additionally, our study found a significant interaction between blood meal source and gonotrophic cycle on fertility. For mammal-fed mosquitoes, fertility was significantly lower in the second gonotrophic cycle compared to the first, a pattern commonly reported in other studies where the first and second cycles are generally the most productive (Awahmukalah & Brooks, 1985; Christiansen-Jucht et al., 2015). This declining fertility pattern has been documented in Cx. quinquefasciatus fed on both birds and mammals (Richards et al., 2012; Telang & Skinner, 2019). However, our results showed that this decline in fertility was not present in bird-fed mosquitoes, indicating a potential role of blood meal source in moderating reproductive output across gonotrophic cycles.

We also investigated whether hatchability could account for the observed variations in fertility among different treatments. Our results indicate no difference in hatchability among treatments.

While our findings corroborate an interaction effect between blood meal source and seasonality, they are not without caveats, as same as other studies. First, our research was conducted under controlled conditions in a laboratory incubator, which implies that our results may differ from those of field studies. Some of the main limitations of this approach include the use of artificial, constant light and temperature rather than a fluctuating environment, and the use of constant humidity, which could affect mosquito performance. However, this is the first study of its kind, and future investigations should aim for more realistic setups. A second limitation, commonly encountered in ecological and behavioral studies, is determining an appropriate sample size for each treatment, as this can impact the ability to detect desired effects or interactions (Taborsky, 2010). To address this issue, we conducted an a priori power analysis to determine the minimum required sample size and repeated the assay multiple times (three replicates). Furthermore, the presence of interaction effects among our explanatory variables suggests that our chosen sample size may have been adequate. Lastly, another aspect of this research that warrants further discussion is the underlying hypothesis that Culex mosquitoes change their feeding habits from birds to mammals. Given that most species are ornithophagic, why is this shift not directed from one bird host to another? In fact, both patterns—bird-to-mammal (Burkett-Cadena et al., 2011) and bird-to-bird (Kent et al., 2009) shifts—are mentioned in the literature. Future studies are aimed to clarify the complex feeding behaviour of Culex mosquitoes regarding seasonal switching, which may depend on several factors, such as habitat, as evidenced in U.S. Culex populations (Caranci, 2010).

In summary, our study investigates the critical feature of host shifting patterns in Culex mosquitoes, which holds significant importance due to their role as bridge vectors between bird and human hosts. Although our study detects a significant interaction between blood meal source and seasonality, we observed the opposite expected trend, suggesting a complex interplay of factors affecting mosquito reproduction and its relationship with host use. We could consider three possible explanations for our results. Firstly, it is possible that additional factors, such as genetic mechanisms or variations in midgut microbiota composition play a role in influencing host use patterns. Secondly, host switching might be explained by factors related to vector biology other than fitness. Lastly, it remains a possibility that Argentinian Cx. quinquefasciatus populations do not exhibit the same host-switching behaviour observed in US populations, and the observed interaction effects may be linked to different mosquito behaviours. However, Spinsanti et al. (2008) found a seasonal variation in WNV human cases, and Beranek (2019) also observed that in autumn, the proportion of blood meals was higher than in summer for Culex quinquefasciatus. Therefore, more in-field studies are needed to assess seasonal host variation in Culex mosquitoes from Argentina.

Further investigations are warranted to corroborate our findings and gain a deeper understanding of the interaction between mosquito feeding pattern and seasonality. Last, these investigations will provide valuable insights into mosquito-borne disease transmission and forecasting its emergence.

Supporting information

eLife Assessment

This useful study provides the first assessment of potentially interactive effects of seasonality and blood source on mosquito fitness, together in one study. During revision, the manuscript has been improved, providing additional solid data to support the robustness of observations. However, the discussion still requires further refinement to present the conclusions in manner that is consistent with the data presented. Overall, this interesting study will advance our current understanding of mosquito biology.

Reviewer #1 (Public review):

Summary:

This study examines the role of host blood meal source, temperature, and photoperiod on the reproductive traits of Cx. quinquefasciatus, an important vector of numerous pathogens of medical importance. The host use pattern of Cx. quinquefasciatus is interesting in that it feeds on birds during spring and shifts to feeding on mammals towards fall. Various hypotheses have been proposed to explain the seasonal shift in host use in this species but have provided limited evidence. This study examines whether the shifting of host classes from birds to mammals towards autumn offers any reproductive advantages to Cx. quinquefasciatus in terms of enhanced fecundity, fertility, and hatchability of the offspring. The authors found no evidence of this, suggesting that alternate mechanisms may drive the seasonal shift in host use in Cx. quinquefasciatus.

Strengths:

Host blood meal source, temperature, and photoperiod were all examined together.

Weaknesses:

The study was conducted in laboratory conditions with a local population of Cx. quinquefasciatus from Argentina. I'm not sure if there is any evidence for a seasonal shift in the host use pattern in Cx. quinquefasciatus populations from the southern latitudes.

Comments on the revision:

Overall, the manuscript is much improved. However, the introduction and parts of the discussion that talk about addressing the question of seasonal shift in host use pattern of Cx. quin are still way too strong and must be toned down. There is no strong evidence to show this host shift in Argentinian mosquito populations. Therefore, it is just misleading. I suggest removing all this and sticking to discussing only the effects of blood meal source and seasonality on the reproductive outcomes of Cx. quin.

Reviewer #2 (Public review):

Summary:

Conceptually, this study is interesting and is the first attempt to account for the potentially interactive effects of seasonality and blood source on mosquito fitness, which the authors frame as a possible explanation for previously observed host-switching of Culex quinquefasciatus from birds to mammals in the fall. The authors hypothesize that if changes in fitness by blood source change between seasons, higher fitness on birds in the summer and on mammals in the autumn could drive observed host switching. To test this, the authors fed individuals from a colony of Cx. quinquefasciatus on chickens (bird model) and mice (mammal model) and subjected each of these two groups to two different environmental conditions reflecting the high and low temperatures and photoperiod experienced in summer and autumn in Córdoba, Argentina (aka seasonality). They measured fecundity, fertility, and hatchability over two gonotrophic cycles. The authors then used generalized linear mixed models to evaluate the impact of host species, seasonality, and gonotrophic cycle on fecundity, fertility, and hatchability. The authors were trying to test their hypothesis by determining whether there was an interactive effect of season and host species on mosquito fitness. This is an interesting hypothesis; if it had been supported, it would provide support for a new mechanism driving host switching. While the authors did report an interactive impact of seasonality and host species, the directionality of the effect was the opposite from that hypothesized. The authors have done a very good job of addressing many of the reviewer's concerns, especially by adding two additional replicates. Several minor concerns remain, especially regarding unclear statements in the discussion.

Strengths:

(1) Using a combination of laboratory feedings and incubators to simulate seasonal environmental conditions is a good, controlled way to assess the potentially interactive impact of host species and seasonality on the fitness of Culex quinquefasciatus in the lab.

(2) The driving hypothesis is an interesting and creative way to think about a potential driver of host switching observed in the field.

Weaknesses:

(1) The methods would be improved by some additional details. For example, clarifying the number of generations for which mosquitoes were maintained in colony (which was changed from 20 to several) and whether replicates were conducted at different time points.

(2) The statistical analysis requires some additional explanation. For example, you suggest that the power analysis was conducted a priori, but this was not mentioned in your first two drafts, so I wonder if it was actually conducted after the first replicate. It would be helpful to include further detail, such as how the parameters were estimated. Also, it would be helpful to clarify why replicate was included as a random effect for fecundity and fertility but as a fixed effect for hatchability. This might explain why there were no significant differences for hatchability given that you were estimating for more parameters.

(3) A number of statements in the discussion are not clear. For example, what do you mean by a mixed perspective in the first paragraph? Also, why is the expectation mentioned in the second paragraph different from the hypothesis you described in your introduction?

(4) According to eLife policy, data must be made freely available (not just upon request).

Author response:

The following is the authors’ response to the previous reviews

We would like to thank you for your valuable comments and suggestions, which have greatly contributed to improving our manuscript.

We have carefully addressed all the reviewers' suggestions, and detailed responses for each Reviewer are provided at the end of this letter. In summary:

• The Introduction has been revised to provide a more focused discussion on results, toning down the speculative discussion on seasonal host shifts.

• The methodology section has been clarified, particularly the power analysis, which now includes a clearer explanation. The random effects in the models have been better described to ensure transparency.

• The Results section was reorganized to highlight the key findings more effectively.

• The Discussion has been restructured for clarity and conciseness, ensuring the interpretation of the results is clearer and better aligned with the study objectives.

• Minor edits throughout the manuscript were made to improve readability and accuracy.

We hope you find this revised version of the manuscript satisfactory.

Reviewer #1 (Public review):

Summary:

This study examines the role of host blood meal source, temperature, and photoperiod on the reproductive traits of Cx. quinquefasciatus, an important vector of numerous pathogens of medical importance. The host use pattern of Cx. quinquefasciatus is interesting in that it feeds on birds during spring and shifts to feeding on mammals towards fall. Various hypotheses have been proposed to explain the seasonal shift in host use in this species but have provided limited evidence. This study examines whether the shifting of host classes from birds to mammals towards autumn offers any reproductive advantages to Cx.

quinquefasciatus in terms of enhanced fecundity, fertility, and hatchability of the offspring. The authors found no evidence of this, suggesting that alternate mechanisms may drive the seasonal shift in host use in Cx. quinquefasciatus.

Strengths:

Host blood meal source, temperature, and photoperiod were all examined together.

Weaknesses:

The study was conducted in laboratory conditions with a local population of Cx. quinquefasciatus from Argentina. I'm not sure if there is any evidence for a seasonal shift in the host use pattern in Cx. quinquefasciatus populations from the southern latitudes.

Comments on the revision:

Overall, the manuscript is much improved. However, the introduction and parts of the discussion that talk about addressing the question of seasonal shift in host use pattern of Cx. quin are still way too strong and must be toned down. There is no strong evidence to show this host shift in Argentinian mosquito populations. Therefore, it is just misleading. I suggest removing all this and sticking to discussing only the effects of blood meal source and seasonality on the reproductive outcomes of Cx. quin.

Introduction and discussion have been modified, toned down and sticked to discuss the results as suggested.

Reviewer #1 (Recommendations for the authors):

Some more minor comments are mentioned below.

Line 51: Because 'of' this,

Changed as suggested.

Line 56: specialists 'or' generalists

Changed as suggested.

Line 56: primarily

Changed as suggested.

Line 98: Because 'of' this,

Changed as suggested.

Reviewer #2 (Public review):

Summary:

Conceptually, this study is interesting and is the first attempt to account for the potentially interactive effects of seasonality and blood source on mosquito fitness, which the authors frame as a possible explanation for previously observed hostswitching of Culex quinquefasciatus from birds to mammals in the fall. The authors hypothesize that if changes in fitness by blood source change between seasons, higher fitness on birds in the summer and on mammals in the autumn could drive observed host switching. To test this, the authors fed individuals from a colony of Cx. quinquefasciatus on chickens (bird model) and mice (mammal model) and subjected each of these two groups to two different environmental conditions reflecting the high and low temperatures and photoperiod experienced in summer and autumn in Córdoba, Argentina (aka seasonality). They measured fecundity, fertility, and hatchability over two gonotrophic cycles. The authors then used generalized linear mixed models to evaluate the impact of host species, seasonality, and gonotrophic cycle on fecundity, fertility, and hatchability. The authors were trying to test their hypothesis by determining whether there was an interactive effect of season and host species on mosquito fitness. This is an interesting hypothesis; if it had been supported, it would provide support for a new mechanism driving host switching. While the authors did report an interactive impact of seasonality and host species, the directionality of the effect was the opposite from that hypothesized. The authors have done a very good job of addressing many of the reviewer's concerns, especially by adding two additional replicates. Several minor concerns remain, especially regarding unclear statements in the discussion.

Strengths:

(1) Using a combination of laboratory feedings and incubators to simulate seasonal environmental conditions is a good, controlled way to assess the potentially interactive impact of host species and seasonality on the fitness of Culex quinquefasciatus in the lab.

(2) The driving hypothesis is an interesting and creative way to think about a potential driver of host switching observed in the field.

Weaknesses:

(1) The methods would be improved by some additional details. For example, clarifying the number of generations for which mosquitoes were maintained in colony (which was changed from 20 to several) and whether replicates were conducted at different time points.

Changed as suggested.

(2) The statistical analysis requires some additional explanation. For example, you suggest that the power analysis was conducted a priori, but this was not mentioned in your first two drafts, so I wonder if it was actually conducted after the first replicate. It would be helpful to include further detail, such as how the parameters were estimated. Also, it would be helpful to clarify why replicate was included as a random effect for fecundity and fertility but as a fixed effect for hatchability. This might explain why there were no significant differences for hatchability given that you were estimating for more parameters.

The power analysis was conducted a posteriori, as you correctly inferred. While I did not indicate that it was performed a priori, you are right in noting that this was not explicitly mentioned. As you suggested, the methodology for the power analysis has been revised to clarify any potential doubts.

Regarding the model for hatchability, a model without a random effect variable was used, as all attempts to fit models with random effects resulted in poor validation. These points have now been clarified and explained in the corresponding section.

(3) A number of statements in the discussion are not clear. For example, what do you mean by a mixed perspective in the first paragraph? Also, why is the expectation mentioned in the second paragraph different from the hypothesis you described in your introduction?

Changed as suggested.

(4) According to eLife policy, data must be made freely available (not just upon request).

Data and code will be publicly available. The corresponding section was modified.

Reviewer #2 (Recommendations for the authors):

Your manuscript is much improved by the inclusion of two additional replicates! The results are much more robust when we can see that the trends that you report are replicable across 3 iterations of the experiment. Congratulations on a greatly improved study and paper! I have several minor concerns and suggestions, listed below:

38-39: I think it is clearer to say "no statistically significant effect of season on hatchability of eggs" ... or specify if you are referring to blood or the interaction of blood and season. It isn't clear which treatment you are referring to here.

Changed as suggested.

54-57: This could be stated more succinctly. Instead of citing papers that deal with specific examples of patterns, I would suggest citing a review paper that defines these terms.

Changed as suggested.

83-84: What if another migratory bird is the preferred host in Argentina? I would state this more cautiously (e.g. "may not be applicable...").

Changed as suggested.

95-96: I don't understand what you mean by this. These hypotheses are specifically meant to understand mosquitoes that DO have a distinct seasonal phenology, so I'm not sure why this caveat is relevant. And naturally this hypothesis is host dependent, since it is based on specific host reproductive investments. I think that the strongest caveat to this hypothesis is simply that it hasn't been proven.

Changed as suggested.

97-115: This is a great paragraph! Very clear and compelling.

Thanks for your words!

118: Do you have an exact or estimated number of rafts collected?

Sorry, I have not the exact number of rafts, but it was at leas more than 20-30.

135: "over twenty" was changed to "several"; several would imply about 3 generations, so this is misleading. If the colony was actually maintained for over twenty generations, then you should keep that wording.

Changed as suggested.

163-164: Can you please clarify whether the replicates were conducted a separate time points?

Changed as suggested.

Note: the track changes did not capture all of the changes made; e.g. 163-164 should show as new text but does not.

You are absolutely right; when I uploaded the last version, I unfortunately deleted all tracked changes and cannot recover them. In this new version, I will ensure that all minimal changes are included as tracked changes.

186 - 189: the terms should be "fixed effect" and "random effect"

Changed as suggested.

191: Edit: linear

Changed as suggested.

194: why was replicate not included as a random effect here when it was above? Also, can you please clarify "interaction effects"? Which interactions did you include?

Changed as suggested. Explained above and in methodology. Hatchability models with random effect variable were poor fitted and validated. The interactions for hatchability were a four-way (season, blood source, cycle and replicate)

207-208: I'm not sure what you mean by "aimed to achieve"? Weren't you doing this after you conducted the experiments, so wouldn't this be determining the power of your model (post-hoc power analysis)? Also, I think you should provide the parameter estimates that were used (e.g. effect size - did you use the effect size you estimated across the 3 replicates?).

Changed as suggested.

214-215: this should be reworded to acknowledge that this is estimated for the given effect size; for example, something like "This sample size was sufficient to detect the observed effect with a statistical power of 0.8" or something along those lines (unless I am misunderstanding how you conducted this test).

Changed as suggested.

246. Abbreviate Culex

Changed as suggested.

253-255: This sentence isn't clear. What do you mean by mixed? Also, the season really seemed to mainly impact the fitness of mosquitoes fed on mouse blood and here the way it is phrased seems to indicate that season has an impact on the fitness of those fed with chicken blood.

Changed as suggested.

258-260: You stated your hypothesis as the relative fitness shifting between seasons, but this statement about the expectation is different from your hypothesis stated earlier. Please clarify.

You are right. Thank you for noting this. It was changed as suggested.

263-266: I also don't understand this sentence; what does the first half of the sentence have to do with the second?

Changed as suggested.

269-270: This doesn't align with your observation exactly; you say first AND second are generally most productive, but you observed a drop in the second. Please clarify this.

Changed as suggested.

280: I suggest removing "as same as other studies"; your caveats are distinct because your experimental design was unique

Changed as suggested.

287: you shouldn't be looking for a "desired" effect; I suggest removing this word

Changed as suggested.

288: It wasn't really a priori though, since you conducted it after your first replicate (unless you didn't use the results from the first replicate you reported in the original drafts?)

It was a posteriori. Changed as suggested.

290: Why is 290 written here?

It was a mistype. Deleted as suggested.

291-298: The meaning of this section of your paragraph is not clear.

Improve as suggested.

304-313: This list of 3 explanations are directed at different underlying questions. Explanations 1 and 2 are alternative explanations for why host switching occurs if not due to differences in fitness. This isn't really an explanation of your results so much as alternative explanations for a previously reported phenomenon. And the third is an explanation for why you may not have observed the expected effect. I suggest restructuring this to include the fact that Argentinian quinqs may not host switch as part of your previous list of caveats. Then you can include your two alternative explanations for host switching as a possible future direction (although I would say that it is really just one explanation because "vector biology" is too broad of a statement to be testable). Also, you haven't discussed possible explanations for your actual result, which showed that mosquito fitness decreased when feeding on mouse blood in autumn conditions and in the second gonotrophic, while those that fed on chicken did not experience these changes. Why might that be?

The discussion was restructured to include all these suggested changes. Additionally, it was also discussed some possible explanations of our results.

315-317: This statement is vague without a direct explanation of how this will provide insight. I suggest removing or providing an explanation of how this provides insight to transmission and forecasting.

Changed as suggested.

319-320: According to eLife policy, all data should be publicly available. From guidelines: "Media Policy FAQs Data Availability Purpose and General Principles To maintain high standards of research reproducibility, and to promote the reuse of new findings, eLife requires all data associated with an article to be made freely and widely available. These must be in the most useful formats and according to the relevant reporting standards, unless there are compelling legal or ethical reasons to restrict access. The provision of data should comply with FAIR principles (Findable, Accessible, Interoperable, Reusable). Specifically, authors must make all original data used to support the claims of the paper, or that is required to reproduce them, available in the manuscript text, tables, figures or supplementary materials, or at a trusted digital repository (the latter is recommended). This must include all variables, treatment conditions, and observations described in the manuscript. The authors must also provide a full account of the materials and procedures used to collect, pre-process, clean, generate and analyze the data that would enable it to be independently reproduced by other researchers."

- so you need to make your data available online; I also understand the last sentence to indicate that code should be made available.

Data and code will be publicly available.

Table 1: it is notable that in replicate 2, the autumn:mouse:gonotrophic cycle II fecundity and fertility are actually higher than in the summer, which is the opposite of reps 1 and 3 and the overall effect you reported from the model. This might be worth mentioning in the discussion.

Mentioned in the discussion as suggested.

Tables 1 and 2: shouldn't this just be 8 treatments? You included replicate as a random effect, so it isn't really a separate set of treatments.

This table reflects the output of the whole experiment, that is why it is present the 24 expetiments.

Figure 3: Can you please clarify if this is showing raw data?

Changed as suggested.

Note: grammatical copy editing would be beneficial throughout

Grammar was improved as suggested.

Version 4

Abstract

Abstract

Host selection by mosquitoes is a keystone to understanding viral circulation and predicting future infection outbreaks. Culex mosquitoes frequently feed on birds during spring and early summer, shifting into mammals towards late summer and autumn. This host switch may be due to changes in mosquito fitness. The aim of this study was to assess if the interaction effect of blood meal source and seasonality may influence reproductive traits of Culex quinquefasciatus mosquitoes. For this purpose, Cx. quinquefasciatus mosquitoes were reared in simulated summer and autumn conditions and fed on two different hosts, chickens and mice, in a factorial design. Fecundity, fertility, and hatchability during two consecutive gonotrophic cycles were estimated. We found greater fecundity and fertility for mosquitoes fed upon birds than mammals. Fecundity and fertility did not vary between seasons for chicken-fed mosquitoes, whereas in autumn they decreased for mouse-fed mosquitoes. These traits decreased in the second gonotrophic cycle for mouse-fed mosquitoes, whereas they did not vary between cycles for chicken-fed mosquitoes. There was no statistically significant effect of blood meal source, seasonality or their interaction on hatchability, hence this variable was similar among treatments. Overall, these results indicate a statistically significant interaction effect of blood meal source and seasonality on fecundity and fertility. However, the pattern was opposite in relation to our hypothesis, suggesting that further studies are needed to confirm and expand our knowledge about mosquito biology and its relationship with seasonal host use shifting.

Introduction

The southern house mosquito, Culex quinquefasciatus Say (Diptera: Culicidae), is a worldwide distributed vector of numerous pathogens, including Wuchereria bancrofti Cobbold (Kasili et al., 2009), Dirofilaria immitis Leidy (Labarthe et al., 1998) and avian malaria (Farajollahi et al., 2011). Besides, this mosquito is also one of the main species responsible for the transmission of the arboviruses West Nile (WNV) and St. Louis encephalitis (SLEV) viruses (Farajollahi et al., 2011). Because of this, this species is of critical concern due to its impact on both public and veterinary health.

Since arboviruses are transmitted by the bite of infected mosquitoes between vertebrate hosts, activity of these pathogens is determined, at least in part, by the blood feeding habits of these vectors. According to its host feeding pattern range, Culex mosquitoes can be regarded as specialists or generalists. Specialist species include mosquitoes feeding primarily on mammals, birds, or reptiles and amphibians. Generalists are mosquitoes fed on the most abundant available hosts (Fikrig and Harrington, 2021). Additionally, some species could experience a seasonal shift in their feeding habits, from avian to mammal hosts (Edman and Taylor, 1968; Hancock and Camp, 2022), being of greater concern from an epidemiological perspective given that they can act as bridge vectors in the transmission of some arboviruses, such as SLEV and WNV (Kilpatrick et al., 2006).

West Nile virus (WNV) and St. Louis encephalitis virus (SLEV) are zoonotic pathogens maintained in nature through an enzootic network involving transmission among viremic birds and ornithophagic mosquitoes (Kilpatrick et al., 2006). However, during late summer and early autumn, these viruses increase their activity in humans and other mammals, which has led to repeated epidemics in the US (Kilpatrick et al., 2006) and South America (Spinsanti et al., 2008). Likewise, several Culex species, including Cx. quinquefasciatus, have exhibited a seasonal shift in their feeding behaviour, feeding primarily on birds in spring and early summer and increasing the proportion of mammal blood meals in late summer and autumn (Thiemann et al., 2011). This seasonal shift suggests that viral activity might spill over from birds to mammals as a result of the mosquito host switch (Edman and Taylor, 1968; Kilpatrick et al., 2006).

Two main contrasting hypotheses have been proposed to explain why primarily ornithophagic mosquitoes increase their feeding on mammals later in the season. The migration hypothesis, proposed by Kilpatrick et al. (2006) for Cx. pipiens, suggests that this shift in feeding pattern is related to the abundance of the preferred avian host, the American robin (Turdus migratorius L.). This hypothesis is based on their findings that as robin densities decrease, the proportion of mammalian blood meals in Cx. pipiens populations increase. However, this hypothesis has several limitations: (i) multiple studies have shown that American robins are not always the preferred host for all Cx. pipiens populations (Apperson et al., 2002; Patrican et al., 2007); (ii) in some areas (e.g., Tennessee) with documented seasonal host shifts, American robins are present year-round (Savage et al., 2007); and (iii) this explanation applies only to mosquitoes whose primary host is the American robin, a species found only in the Northern Hemisphere, or another migratory bird. Therefore, it might not be applicable to southern regions, such as Argentina, where WNV seasonality is documented (Spinsanti et al., 2008) but American robin is absent. Because of this, it seems that the hypothesis proposed by Kilpatrick et al. (2006) might be habitat dependent (Caranci, 2010). On the other hand, the defensive behaviour hypothesis proposed by Burkett-Cadena et al. (2011), suggests that host breeding cycles drive the host shift of mosquitoes. According to this hypothesis, in hosts with parental care, during reproductive seasons there is a greater investment of energy in assuring the offspring survivorship, consequently producing an increase in susceptibility of being bitten by mosquitoes as a result of a decrease in defensive behaviours. This event leads to the detection of peaks of host use during periods of reproductive investment, in summer for birds and autumn for mammals (Burkett-Cadena et al., 2011). While this hypothesis was proposed as an alternative to address the limitations of the migration hypothesis, it also has drawbacks. It is based on a limited range of host species, assumes a distinct seasonal phenology—which is not present for all mosquito hosts—and remains untested, leaving its validity uncertain.

Explanations for the seasonal shift in mosquito host feeding focused on vector biology are largely lacking in the literature. Numerous biological factors, such as stress, metabolic rate, and blood meal source, along with environmental variables like temperature and photoperiod, influence mosquito physiology and can affect reproductive traits such as fecundity, development rate, and survivorship. These factors can give rise to new nutritional requirements that may ultimately lead to seasonal variations in host selection by mosquitoes (Ciota et al., 2014; Costanzo et al., 2015; Gervasi et al., 2016; Yan et al., 2018, 2017). While several studies have examined the effects of these biological and environmental variables on mosquito reproduction, the complex nature of mosquito biology suggests that multiple interactions among these variables may produce novel responses that are not apparent when considered individually. The aim of this study was to assess whether there is an interaction effect between the source of the host blood meal and seasonality (defined by temperature and photoperiod) on three reproductive traits of Culex quinquefasciatus mosquitoes: fecundity, fertility, and hatchability. We hypothesize that the interaction between these two variables influences the reproductive outcomes, potentially leading to a seasonal shift in host selection, driven by a reproductive advantage. Given the reported seasonal changes in host use by Cx. quinquefasciatus, in autumn we expect a greater number of eggs (fecundity) and larvae (fertility) in mosquitoes after feeding on a mammal host compared to an avian host, and the opposite trend in summer.

Materials and methods

Establishment and maintenance of mosquitoes

Egg rafts of Culex quinquefasciatus were collected from a drainage ditch at Universidad Nacional de Córdoba Campus, Córdoba city, in February 2021. Each raft was individually maintained in plastic containers with one liter of distilled water. The hatched larvae were fed with 100 mg of liver powder three times per week until pupation. Pupae were then transferred to plastic emerging cages (21 cm x 12 cm) covered with a tulle-like fabric, containing distilled water but no food. Adults emerging from each raft were identified morphologically (Darsie, 1985) and molecularly (Smith and Fonseca, 2004) to ensure they corresponded to Cx. quinquefasciatus, since Cx. pipiens and its hybrids coexist sympatrically in Córdoba city (Branda et al., 2021). All adults were reared in a cardboard cage of 22.5 liters (28 cm x 36.5 cm) and provided ad libitum with a 10% sugar solution soaked in cotton pads placed on plastic cups. For long-term maintenance of the colony, 24-hour-starved mosquitoes were offered a blood meal from a restrained chicken twice a month. Four days after feeding, a plastic container with distilled water was placed inside the cage to allow engorged females to lay egg rafts. Batches of egg were collected and transferred to plastic containers (30 cm x 25 cm x 7 cm) in a proportion of three rafts per container, filled with three liters of distilled water. The hatched larvae were also fed with liver powder at the same proportion described above. Pupae were transferred to emerging cages, and adult mosquitoes were placed in the final cardboard cage. The colony has been maintained for over twenty generations in the Insectary of the Instituto de Virología “Dr. J.M. Vanella” (InViV). Room conditions were controlled at 28 ± 1 °C, with a 12L:12D photoperiod and 70% relative humidity.

Experimental design

Blood source. For experimental trials, avian hosts (live chicks of the species Gallus gallus) and mammalian hosts (live mice of the species Mus musculus, strain C57BL/6) were used to evaluate the effect of blood meal source on fecundity, fertility, and hatchability. Chicks were generously donated by the Bartolucci poultry farm (Córdoba, Argentina), while mice were commercially obtained from the Instituto de Investigación Médica Mercedes y Martín Ferreyra (CONICET - Universidad Nacional de Córdoba). In each trial, 24-hour-starved adult female mosquitoes were provided with a blood meal from restrained chicks or mice. Vertebrate hosts were offered to mosquitoes one hour before the lights were turned off and were kept for 3 hours during 2 consecutive gonotrophic cycles.

The experimental use of animals (mosquitoes and vertebrates) was approved by the ethical committee at Facultad de Ciencias Médicas, Universidad Nacional de Córdoba (FCM-UNC) in compliance with the legislation regarding the use of animals for experimental and other scientific purposes (accession code: CE-2022-00518476-UNC-SCT#FCM).

Seasonality

To assess the effect of seasonality (photoperiod + temperature) on fecundity, fertility, and hatchability, a “typical” summer and autumn day from Córdoba city was simulated in an incubator, where mosquitoes were housed throughout the entire experiment. Summer conditions were as follows: T°min = 22°C, T°max = 28°C, photoperiod = 14L:10D. Autumn conditions were characterized by: T°min = 16°C, T°max = 22°C, and photoperiod = 10L:14D. The humidity levels were maintained at 60-70% for both conditions. Data for simulated conditions were obtained from Climate Data website (https://es.climate-data.org/).

Feeding trial design

The interaction effect of blood source (chicken or mouse) and seasonality (autumn or summer) was evaluated during two consecutive gonotrophic cycles (I and II). This combination produced eight colonies that were established using egg rafts collected from the maintenance colony: mouse:autumn-I, mouse:autumn-II, chicken:autumn-I, chicken:autumn-II, mouse:summer-I, mouse:summer-II, chicken:summer-I, chicken:summer-II. This set of eight colonies was repeated in triplicate (1-3) at different time points, generating a total of 24 treatments.

The experimental colonies were maintained under controlled conditions to ensure that all adults were of the same size (Kauffman et al., 2017). Adult mosquitoes were placed in cardboard cages of 8.3 liters (21 cm x 24 cm) and were provided with ad libitum access to a 10% sugar solution, which was soaked in cotton pads placed on plastic cups.

Feeding trials were conducted at two time points: 5 days post-emergence (first cycle) and 14 days post-emergence (second cycle). Following each blood meal, female mosquitoes were anesthetized using CO2 and classified as fully engorged (VI Sella’s stage), partially fed (I-V Sella’s stage), or unfed (I Sella’s stage), following the classification by Santos et al. (2019). Fully engorged females were counted and separated into a separate cage to complete their gonotrophic cycle. Four days after feeding, a cup containing distilled water was placed inside the cage to facilitate oviposition. After each oviposition cycle, egg rafts were collected, counted, and transferred to a 12-well plate containing 4 mL of distilled water. They were then photographed to subsequently determine the number of eggs per raft. Rafts were maintained in each well until they hatched into L1 larvae, at which point they were preserved with 2 mL of 96% ethanol, and the number of L1 larvae per raft was counted.

Statistical analysis

The reproductive outputs measured for each experimental colony were fecundity, fertility, and hatchability. All analyses were performed using R Studio statistic software, v.4.2.1 (R Core Team, 2022).

Fecundity was defined as the number of eggs per raft and fertility as the number of L1 larvae hatched per raft. To evaluate the effect of blood meal source and seasonality on these two variables, a Generalized Linear Mixed Model (GLMM; lme4 package) with negative binomial error distribution and logarithmic link function was adjusted (Bates et al., 2015). Fecundity and fertility served as response variables, while the fixed effect variables included blood source, seasonality, and gonotrophic cycle, all in a three-way interaction. Treatment (n = 24) was set as a random effect variable.

Hatchability (= hatching rate) was defined as the ratio of the number of larvae to the number of eggs per raft. While the original intent was to apply a GLMM to all three variables, the validation of GLMMs for hatchability across several distributions revealed poor fit. Consequently, a Generalized Linear Model (GLM; MASS package) with a quasipoisson error distribution and logarithmic link function was used instead (Venables and Ripley, 2002). In this model, the response variable was the number of larvae, while the explanatory variables included blood source, seasonality, gonotrophic cycle, and replicate as a four-way interaction. Additionally, the logarithm of the number of eggs was incorporated as an offset in the model.

Multiple comparisons among treatments were conducted using the Tukey’s honestly significant difference test (HSD Tukey), incorporating the Kramer (1956) correction for unbalanced data, also known as the Tukey-Kramer method. This approach allows to set the family-wise error rate to 0.05. These comparisons were performed using the emmeans package (Lenth, 2017).

Model validation was assessed using a simulation-based approach to generate randomized quantile residuals to test goodness of fit (Q-Q plot), homoscedasticity, and outliers and influential points (Cook’s distance and Leverage plots). These analyses were conducted with the DHARMa package (Hartig, 2024).

Additionally, post-hoc statistical power was evaluated using the mixedpower package, a tool specifically designed for simulating statistical power in Generalized Linear Mixed Models (GLMMs). The analysis proceeded in three key steps: (1) simulation of new datasets based on the parameters of an original fitted model (the fecundity and fertility models), (2) refitting the model to each dataset and testing for statistical significance, and (3) calculating statistical power as the proportion of simulated datasets where the effect of interest reached significance. Fixed effects were simulated for sample sizes ranging from 30 to 150, with 1,000 datasets generated at each sample size to ensure robust estimation. The variable treatment was designated as a random effect, capturing variability associated with this factor. For further details on the methodology, including considerations for simulation accuracy and model structure, refer to Kumle et al. (2021).

Results

A total of 1,162 egg rafts were obtained for analysis, distributed across the eight treatments as follows: chicken:autumn-I: 260, chicken:autumn-II: 174, chicken:summer-I: 162, chicken:summer-II: 88, mouse:autumn-I: 146, mouse:autumn-II: 96, mouse:summer-I: 167, and mouse:summer-II: 69. The full count of egg rafts for each replicate (24 treatments), along with the means for fecundity, fertility, and hatchability, are summarized in Table 1. This sample size was sufficient to detect the observed effect with a statistical power of 0.8, indicating an 80% chance of detecting a true effect if it exists (Figure SM1).

In both models—fecundity and fertility—the interaction between blood source and seasonality was statistically significant (fecundity: LRT X² = 5.69, p < 0.05; fertility: LRT X² = 4.37, p < 0.05) (Tables 2,3; Figures 1,2A). Mosquitoes that fed on chicken blood had the highest fecundity (summer: 136.4 eggs/raft, autumn: 141.5 eggs/raft) and fertility (summer: 123.9 larvae/raft, autumn: 126 larvae/raft) in both seasons. In summer, fecundity and fertility were 7% and 11% higher, respectively, in chicken-fed mosquitoes compared to mouse-fed mosquitoes. In autumn, these differences increased to 46% for fecundity and 52% for fertility in favor of chicken-fed mosquitoes. No significant seasonal variation was observed in fecundity and fertility for chicken-fed mosquitoes, but both measures were statistically significant lower by 24% and 25%, respectively, in autumn compared to summer for mouse-fed mosquitoes.

Interaction plot showing the effect of blood meal source (chicken or mouse) and seasonality (autumn or summer) on fecundity (eggs/raft). Points represent the predicted marginal mean values and the bars its corresponding confidence intervals. Means sharing same letters are not statistically different (p > 0.05).

Interaction plots showing the effect of blood meal source (chicken or mouse) and seasonality (autumn or summer) (A) or blood meal source and gonotrophic cycle (first or second) (B) on fertility (eggs/raft) of Culex quinquefasciatus. Points represent the predicted marginal mean values and the bars its corresponding confidence intervals. Means sharing same letters are not statistically different (p > 0.05).

In the fertility model, there was also a significant interaction between blood source and gonotrophic cycle (LRT X² = 3.98, p < 0.05), indicating that the impact of blood type on larvae production differed between the first and second cycles (Table 3, Figure 2B). Chicken-fed mosquitoes showed no difference in fertility between cycles, while those fed on mouse blood had 25% lower fertility in the second cycle compared to the first. During the second cycle, fertility was 50% higher in chicken-fed mosquitoes compared to mouse-fed mosquitoes.

Figure 3 presents the predicted mean values of fecundity and fertility for the 24 treatments, with no differences observed across the three replicates. The variance of random intercepts among treatments was approximately 0.006 for fecundity and 0.007 for fertility, accounting for 6% and 4% of the total variance in the respective models.

Dotplots showing the variation of predicted fecundity (eggs/raft) and fertility (larvae/raft) means along the 24 treatments, with combination of blood source (chicken or mouse), seasonality (autumn or summer), gonotrophic cycle (first or second) and replicate (1 to 3).

There was no statistical difference among treatments and replicates in terms of hatchability, given that no statistical effects of blood source, seasonality or gonotrophic cycle were observed (Table 4). The mean hatchability for all treatments was 0.89, with a range between 0.84 for the mouse:autumn-II (replicate 1) colony and 0.96 for mouse:autumn-I (replicate 1) (Table 1, Figure 4).

Boxplots showing predicted mean hatchability for the 24 treatments, combining replicate (1-3), blood meal source (chicken or mouse), seasonality (summer or autumn) and gonotrophic cycle (first or second), of Culex quinquefasciatus.

Discussion

Understanding the factors influencing mosquito host selection is crucial for predicting arbovirus transmission dynamics. Seasonal shift in host utilization have been widely attributed to changes in host behaviour (Burkett-Cadena et al., 2011; Kilpatrick et al., 2006), yet the role of mosquito biology in this process remains largely unexplored. Blood meal source and seasonality are known to affect mosquito reproductive success, but their combined effect has received little attention. Our study tested the hypothesis that the interaction between blood meal source and seasonality could affect mosquito fitness, a key factor that could play a role in seasonal host shift. By examining this interaction, we aimed to assess whether physiological processes like reproduction could be considered a candidate for shaping seasonal feeding patterns in Culex quinquefasciatus.

If seasonal host shift were driven by fitness advantages, we would expect mosquitoes to exhibit higher reproductive output when feeding on the predominant host in each season—that is, greater fecundity and fertility with bird blood in summer and with mammal blood in autumn. However, our findings did not fully align with these expectations. While mosquitoes fed on birds maintained high reproductive success in both summer and autumn, those fed on mammals exhibited a significant decline in reproductive output during autumn. This suggests that, rather than conferring a reproductive advantage in this season, mammalian blood may become a suboptimal resource under autumn conditions, potentially due to metabolic constraints or differences in nutrient assimilation.

The effect of blood meal on fitness observed in this study (greater in bird-fed mosquitoes) is consistent with previous findings on Cx. quinquefasciatus (Richards et al., 2012; Telang and Skinner, 2019) and other Culex (Demirci et al., 2014) and Aedes (Harrison et al., 2021) species. Avian blood is known to be nutritionally richer than mammalian blood, particularly in proteins and essential amino acids (e.g., isoleucine) required for egg development (Alto et al., 2014). Beyond intrinsic differences in blood composition, environmental factors such as temperature and photoperiod (seasonality) may further modulate how nutrients from a blood meal are allocated. While most of the nutrients obtained from a blood meal are directed toward egg production, a portion is also required for non-reproductive physiological functions (Rivera-Pérez et al., 2017), and the extent to which these resources are utilized may depend on external conditions such as temperature (Briegel and Timmermann, 2001).

Previous studies have shown that temperature regulates oogenesis and influences how mosquitoes allocate proteins obtained from a blood meal. In Aedes aegypti, for example, the proportion of blood-derived proteins allocated to egg production varies with temperature: at low temperatures, approximately 50% of these proteins are used for oogenesis, whereas at higher temperatures, this proportion drops to 25% (Briegel and Timmermann, 2001). If a similar mechanism occurs in Cx. quinquefasciatus, it could partially explain why mammal-fed mosquitoes exhibited a more substantial reproductive decline in autumn. Under lower temperatures, a greater proportion of proteins may be required for egg development, which could disproportionately impact individuals feeding on nutritionally poorer blood sources.

In addition to this blood meal-seasonal effect, we also found a significant interaction between blood meal source and gonotrophic cycle. This result suggests that the impact of blood sources is not uniform across cycles. If blood meal source were the only determinant of reproductive success, we would expect its effects to be consistent across cycles, with mosquitoes fed on birds maintaining higher reproductive output than those fed on mammals in a stable pattern. However, our findings showed that while bird-fed mosquitoes sustained similar reproductive output across cycles, those fed on mammals exhibited a significant decline in fertility in the second cycle. Since each cycle involves a new blood meal, this suggests that factors beyond blood nutritional quality contribute to the observed pattern, disproportionately affecting mosquitoes feeding on mammals in later cycles.

A decline in fertility across gonotrophic cycles has also been documented in other Cx. quinquefasciatus populations fed on both birds and mammals (Richards et al., 2012; Telang and Skinner, 2019), suggesting that reproductive output varies not only with seasonality but also with the number of successive cycles. One potential explanation for this decline is age-related reproductive constraints. With each successive cycle, ovarian follicular degeneration increases, leading to a reduction in the number of viable eggs per raft (Awahmukalah and Brooks, 1985). Additionally, given that several days separate the first and second cycles, physiological aging could contribute to this pattern (McCann et al., 2009). Interestingly, bird-fed mosquitoes maintained stable fertility across cycles, a pattern that has also been observed in previous studies (Bennett, 1970). While fertility often declines in later cycles, some studies have reported no change or even an increase in egg production under certain conditions (Bennett, 1970). This variability suggests that the effects of successive gonotrophic cycles on fertility may depend on multiple factors, including blood meal quality. Since avian blood is known to be nutritionally richer than mammalian blood, mosquitoes feeding on birds may have access to sufficient resources to sustain egg production across cycles, even as they age. However, further research is needed to determine whether specific nutrient components play a role in maintaining reproductive output over time. Similarly, body size has been shown to influence reproductive patterns in some mosquito species. In Aedes aegypti, for instance, larger females exhibit a rapid decline in fecundity after the first cycle, whereas smaller females maintain consistent egg production across early cycles before experiencing a significant drop (Briegel et al., 2002). While we did not assess body size in this study, these findings underscore the complexity of reproductive trade-offs in mosquitoes and the importance of considering multiple interacting factors when examining reproductive dynamics across cycles.

Together, these results indicate that both blood meal source and seasonality influence mosquito reproductive output. Moreover, hatchability did not differ among treatments, indicating that differences in egg viability did not drive these effects. This reinforces the idea that the main differences in reproductive output stem from factors affecting fecundity and fertility rather than post-oviposition processes.

While our findings confirm an interaction effect between blood meal source and seasonality, several caveats must be considered. First, our study was conducted under controlled laboratory conditions, which may not fully capture the environmental variability experienced by mosquitoes in the field. Factors such as fluctuating temperature, natural light cycles, and variable humidity levels could influence mosquito reproductive performance in ways not accounted for in our setup. However, as this is the first study of its kind, future research should aim to incorporate more ecologically realistic conditions. A second limitation commonly encountered in ecological and behavioural studies, is determining an appropriate sample size for detecting biologically meaningful effects (Taborsky, 2010). To address this, we repeated the assay across three replicates and conducted a post-hoc power analysis, which suggested that our sample size was adequate given the presence of interaction effects among explanatory variables. Finally, it remains possible that Cx. quinquefasciatus populations in Argentina do not exhibit the same host-switching behaviour observed in US populations, potentially influencing the relevance of our findings for different ecological contexts. However, previous studies have reported seasonal shifts in host selection and arbovirus transmission patterns in Argentina (Beranek, 2019; Spinsanti et al., 2008), suggesting that similar dynamics may be at play. More field-based studies will be essential to assess seasonal host variation in Cx. quinquefasciatus and its implications for arbovirus transmission.

In summary, our study investigates a critical aspect of mosquito reproductive success and its potential link to seasonal host selection in Culex mosquitoes, a group of significant epidemiological importance due to their role as bridge vectors between avian and human hosts. While we detected a significant interaction between blood meal source and seasonality, our findings do not support the idea that reproductive fitness alone explains seasonal host shifts. Instead, they suggest that the relationship between mosquito reproduction and host selection is more complex, likely influenced by multiple ecological and physiological factors beyond fitness constraints. Given this complexity, alternative explanations for seasonal host shifts should be considered. Genetic predisposition, midgut microbiota composition, physiological constraints, or other biological mechanisms unrelated to fitness may also contribute to shaping host-use patterns (Gervasi et al., 2016; Yan et al., 2018). Investigating these aspects will be crucial to fully understand the drivers of host selection in mosquitoes. Future research should aim to unravel the intricate relationship between feeding patterns, reproductive physiology, and environmental conditions. Refining our understanding of these dynamics will improve vector ecology models and enhance predictions of how seasonal variations in host selection may influence pathogen transmission.